---
title: "age_standardize()"
format:
  html:
    embed-resources: true
  gfm:
    prefer-html: false
vignette: >
  %\VignetteIndexEntry{age_standardize()}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{knitr}
editor: source
---

```{r}
#| echo: false
#| message: false
library(rads)
```

```{r}
#| echo: false
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

## Introduction

The comparison of crude health indicator rates across populations is often confounded by different age structures. For example, in 2019 the mean age of Hispanic King County residents was \~12.5 years younger than that of white residents (27.5 vs 40). Given this age discrepancy, we should expect higher mortality rates among white residents. However, you might be thinking, *Is there a way to compare Hispanic and White mortality rates while fully accounting for their different age structures? I.e., is there a way to standardize or adjust for age?*

While you can use Poisson regression or other advanced statistical techniques to calculate age adjusted rates, for simplicity and broader comparison with the work of others, it is often beneficial to adjust your data using a standard reference population. This process is called ***age standardization***. Age standardization answers the question, *What rates would I observe if my data had the same age structure as a different population?*

There are 35 reference populations baked into `rads` (via its dependency on [`rads.data`](https://github.com/PHSKC-APDE/rads.data)). Their names are visible when you type `list_ref_pop()`. Here are the first five as an example:

```{r}
list_ref_pop()[1:5]
```

The first one is the default reference population used by WA DOH and the `rads` package. Taking a peek behind the curtain shows that each reference standard is a simple table of age ranges with their corresponding populations.

```{r}
#| echo: true
#| results: 'hide'
get_ref_pop("2000 U.S. Std Population (11 age groups)")[, 1:4]
```

```{r}
#| echo: false
pretty_kable(get_ref_pop("2000 U.S. Std Population (11 age groups)")[, 1:4])
```

The `age_standardize()` function provides a simple way to use reference populations to calculate age standardized rates. This vignette will provide some examples of how to use `age_standardize()` and how to understand its output. To get the most out of this vignette, we recommend that you type each and every bit of code into R. Doing so will almost definitely help you learn the syntax much faster than just reading the vignette or copying and pasting the code.

## `age_standardize()` arguments

Arguments are the values that we send to a function when we call it. The standard arguments for `age_standardize()` are:

1)  `ph.data`: a data.table or data.frame containing the data to be age-standardized. At a minimum, it must include the following columns:

    - an integer column named 'age' representing individual ages, OR a column named 'agecat' containing pre-aggregated age categories that match the bins in the reference population
    - a numeric column of event counts (see `my.count` below)
    - a numeric column of population counts (see `my.pop` below)

2)  `ref.popname`: a character vector whose only valid options are "none" (used when ph.data has a 'stdpop' column) or one of the reference standards contained in `list_ref_pop()`

3)  `my.count`: a character vector, identifies the column name for the event count data in ph.data. The default is "count"

4)  `my.pop`: a character vector, identifies the column name for the population denominator corresponding to the aggregated count. The default is "pop"

5)  `collapse`: a logical vector (T\|F), determines whether or not to collapse the 'age' column in ph.data to match those in ref.popname. If your data is already collapsed / aggregated into the same age bins as your reference population, and if that variable is named "agecat", then you can set collapse to F. The default is T

6) `line_level`: a logical vector (T\|F), requires `collapse = T`. Set to `TRUE` if `ph.data` is line-level (un-aggregated) data, e.g., one row per event with a population total merged onto every row. The default is `FALSE`, in which case `age_standardize()` only warns about un-aggregated data rather than changing how it's summed. See [Example #1](#example-1-the-basics) below.

7)  `per`: an integer, used as multiplier for all rates and CI, e.g., when per = 1000, the rates are per 1000 persons. The default is 100,000

8)  `conf.level`: a numeric value between 0.00 & 0.99, designates the confidence level (i.e., 1 - alpha) to be used in the calculations. The default is 0.95

9)  `by`: a character vector, specifies the variable(s) by which to stratify the rate results (if any). The default is NULL (i.e., no stratification)

10)  `diagnostic_report`: If `by` is used and there are groups with missing ages, setting `diagnostic_report = TRUE` returns a diagnostic table instead of normal results. Use this option if a warning about missing age groups appears when running the function normally.

## Preparing your dataset

Your dataset must have the following three columns ...

1)  'age' or 'agecat':

    - 'age' in single years (if collapse = T)

    - 'agecat' with the same age bins as your selected reference population (if collapse = F)

2)  a count for the event (e.g., disease) for which you want to find an age standardized rate

3)  the population corresponding to the age or agecat in your original data

`ph.data` can have pre-aggregated counts (each unique `age` and `by` combination has a single row) or un-aggregated (line-level) counts. They are equally valid as long as you understand how to use the `collapse` and `line-level` arguments. In short, when you set `collapse = TRUE` & `line_level = TRUE`, the function will aggregate line-level data for you. The examples below should help you understand the syntax. Please see ``?age_standardize`` for more details.

## Note about incomplete ages

If you are using an 'age' column and do not have all ages [0, 100], you will receive a warning.

Incomplete ages are expected and **required** if you are performing analyses for a specific age group (e.g., adolescent birth rates, opioid mortality rate among working age adults, etc.).

Otherwise, to ensure you calculate accurate rates, be sure to include every age from 0 to 100. If there were no events/counts for a given age, set count = 0 and use the corresponding actual population for that age.

------------------------------------------------------------------------

## Example #1: The Basics

**Create synthetic line level data for a cohort 51 to 60 years of age with a binary indicator for disease status**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
library(data.table)
set.seed(98104)

temp1 <- data.table(
  age = c(rep(51:60, 100), 0:50, 61:100),
  disease = c(sample(0:1, 1000, replace = TRUE), rep(0, 91)),
  pop = c(rep(seq(1000, 910, -10), 100), rep(0, 91))
)

head(temp1)
```

```{r}
#| echo: false
pretty_kable(head(temp1))
```

**Run age_standardize directly on the line-level data**

Notice that we have *not* aggregated `temp1`. There are 100 rows for every age, and `pop` is simply the (correct) population total for that age, repeated on every one of those rows. This is a very common way to end up with un-aggregated data by accident: a population total gets merged onto individual-level event records before they've been collapsed to one row per age. 

Please read the two warnings below the following code block so that you become familiar with the function's feedback. 

```{r}
#| warning: true
#| echo: true
ex1.0 <- age_standardize(ph.data = temp1,
                         ref.popname = "2000 U.S. Std Population (11 age groups)",
                         collapse = T,
                         my.count = "disease",
                         my.pop = "pop",
                         per = 1000,
                         conf.level = 0.95)
```

```{r}
#| echo: false
pretty_kable(head(ex1.0))
```

- **count** is the number of events that occured (i.e., the numerator)

- **pop** is the population corresponding to the count (i.e., the denominator)

- **crude.rate, crude.lci, & crude.uci** are the crude (i.e., observed) rate and CI

- **adj.rate, adj.lci, & adj.uci** are the age standardized rate and CI

*Does anything seem fishy? Think for a moment ...*

The warning above told you: `age_standardize()` noticed that `age` has more than one row per value, but by default it doesn't know whether that's a mistake or not, so it just warns and aggregates (sums) and the `pop` and the `disease`. Since `pop` is really a *single* population total repeated 100 times per age, that naive sum inflates it 100-fold: total `pop` is now `r format(sum(ex1.0$pop), big.mark = ',')`, when it should be closer to 10,000.

**Set `line_level = TRUE` to fix this automatically**

```{r message=TRUE, warning=FALSE}
#| echo: true
ex1.0b <- age_standardize(ph.data = temp1,
                          ref.popname = "2000 U.S. Std Population (11 age groups)",
                          collapse = T,
                          my.count = "disease",
                          my.pop = "pop",
                          per = 1000,
                          conf.level = 0.95,
                          line_level = TRUE)
```

```{r}
#| echo: false
pretty_kable(head(ex1.0b))
```

With `line_level = TRUE`, `age_standardize()` checked whether `pop` is *identical* across the duplicate rows for each age. It summed `disease` within each age and used the unique `pop` value for a given age, and told you so with the message above. The total `pop` (`r format(sum(ex1.0b$pop), big.mark = ',')`) is now reasonable.

*Why isn't this the default?* Because `age` + `by` groups with different `pop` aren't always a mistake. It's also what you get from data that's properly pre-aggregated at a *finer* grain than you're asking `age_standardize()` to report. For example, data aggregated by age **and** sex, summarized *without* stratifying by sex:

```{r}
#| warning: true
#| echo: true
by_sex <- data.table(sex = rep(c("F", "M"), each = 2),
                     age = rep(c(51, 52), 2),
                     disease = c(1, 2, 3, 4),
                     pop = c(480, 500, 520, 510)) # pop differs by sex by age

ex1.0c <- age_standardize(ph.data = by_sex,
                ref.popname = "2000 U.S. Std Population (11 age groups)",
                collapse = T, 
                my.count = "disease", 
                my.pop = "pop")[]
```

```{r}
#| echo: false
pretty_kable(head(ex1.0c))
```

You still get the warning (there *are* multiple rows per age), but `pop` for each age is correctly summed across sexes to get the overall (non-stratified) population. If we ran this with `line_level = TRUE` instead, the result would be identical: `pop` differs across the duplicate rows for each age, so `age_standardize()` sums it rather than deduplicating it, exactly like the default. `line_level = TRUE` only ever changes the result when `pop` is *identical* across duplicate rows.

**You can always choose to pre-aggregate (sum) the line-level data yourself instead**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
temp1 <- temp1[, .(disease = sum(disease)), by = c("age", "pop")]
head(temp1)
```

```{r}
#| echo: false
pretty_kable(head(temp1))
```

**Run age_standardize again \<\< this time with aggregated disease events**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
ex1.0d <- age_standardize(ph.data = temp1,
                          ref.popname = "2000 U.S. Std Population (11 age groups)", 
                          collapse = T,
                          my.count = "disease", 
                          my.pop = "pop", 
                          per = 1000, 
                          conf.level = 0.95)
head(ex1.0d)
```

```{r}
#| echo: false
pretty_kable(head(ex1.0d))
identical(ex1.0b, ex1.0d)
```

```{r}
#| echo: true
identical(ex1.0b, ex1.0d)
```

As expected, `ex1.0d` is identical to `ex1.0b` above. This proves that pre-aggregating by hand and letting `line_level = TRUE` do it automatically produce the same result.

**Run age_standardize again \<\< this time with a different reference population**

Now let's see what happens if we change the reference population. We'll arbitrarily set the reference population to the 19th in the list provided by `list_ref_pop()` (i.e., '`r list_ref_pop()[19]`').

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
ex1.0e <- age_standardize(ph.data = temp1,
                         ref.popname = list_ref_pop()[19], 
                         collapse = T,
                         my.count = "disease", 
                         my.pop = "pop", 
                         per = 1000, 
                         conf.level = 0.95)
head(ex1.0e)
```

```{r}
#| echo: false
pretty_kable(head(ex1.0e))
```

As we'd expect, the crude rates are identical since the reference population is irrelevant for those calculations (`r ex1.0d$crude.rate` vs `r ex1.0e$crude.rate`). However, the age-standardized rates changed substantially from `r ex1.0d$adj.rate` to `r ex1.0e$adj.rate`. Remember, if you want to compare your age-adjusted rates to those published by other health jurisdictions, it's important that you use the same reference population.

------------------------------------------------------------------------

## Example #2: Stratification

**Let's create a new dataset, with counts aggregated by age and gender (F\|M)**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
set.seed(98104)
temp2 <- data.table(
  gender = sample(c('F', 'M'), 20000, replace = T), 
  age = sample(0:100, 20000, replace = T),
  disease = sample(0:1, 20000, replace = T))

# aggregate
temp2 <- temp2[, .(pop = .N, disease = sum(disease)), .(gender, age)]

setorder(temp2, age, gender)

head(temp2)
```

```{r}
#| echo: false
pretty_kable(head(temp2))
```

**Let's examine the overall rates**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
ex2.1 <- age_standardize(ph.data = temp2,
                         collapse = T,
                         my.count = "disease", 
                         my.pop = "pop", 
                         per = 1000, 
                         conf.level = 0.95)

head(ex2.1)
```

```{r}
#| echo: false
pretty_kable(head(ex2.1))
```

Notice that we didn't include the `ref.popname` argument. When it is not specified, `age_standardize()` uses the default which is `list_ref_pop()[1]` (i.e, `r list_ref_pop()[1]`).

**Now let's run the same analysis, but stratified by gender**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
ex2.2 <- age_standardize(ph.data = temp2,
                       collapse = T,
                       my.count = "disease", 
                       my.pop = "pop", 
                       per = 1000, 
                       conf.level = 0.95, 
                       by = "gender")

head(ex2.2)
```

```{r}
#| echo: false
pretty_kable(head(ex2.2))
```

Here we see that the crude rates are nearly identical (`r ex2.2[gender == 'F']$crude.rate` vs `r ex2.2[gender == 'M']$crude.rate`), but the adjusted rate is higher for females (`r ex2.2[gender == 'F']$adj.rate` vs `r ex2.2[gender == 'M']$adj.rate`).

------------------------------------------------------------------------

## Example #3: Using your own standard population

**Create a reference population for the gendered dataset above**

To keep things simple, we will create a reference population based on single ages rather than age bins. As specified in the arguments description above, we will name the standard population column 'stdpop'.

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
set.seed(98104)
new.standard <- unique(temp2[, .(gender, age)])
new.standard[, stdpop := fifelse(gender == 'M', 
                                 sample(7800:16000, 1, replace = T), 
                                 sample(10000:20000, 1, replace = T))]
head(new.standard)
```

```{r}
#| echo: false
pretty_kable(head(new.standard))
```

**Merge the standard population onto the data**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
temp3 <- merge(temp2, 
               new.standard, 
               by = c("age", "gender"), 
               all = T)

head(temp3)
```

```{r}
#| echo: false
pretty_kable(head(temp3))
```

**Calculate the rates when `ref.pop = "none"`**

The data set already contains a `stdpop` column, so we can set `ref.popname = "none"`. Also note that we need to specify `collapse = F` because the function expects data to be pre-aggregated when you have the column `stdpop`.

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
ex3.1 <- age_standardize(ph.data = temp3,
                       ref.popname = "none",
                       collapse = F,
                       my.count = "disease", 
                       my.pop = "pop", 
                       per = 1000, 
                       conf.level = 0.95, 
                       by = "gender")

head(ex3.1)
```

```{r}
#| warning: false
#| message: false
#| echo: false
pretty_kable(head(ex3.1))
```

Notice that the crude rates are identical to those in ex2.2, but the adjusted rates differ due to the use of our custom reference population.

------------------------------------------------------------------------

## Example #4: When else should I specify `collapse = F`?

In example #3 above, we specified `collapse = F` because `age_standardize()` expects the data to be pre-aggregated when you provide a `stdpop` column. The other time when you will want to set `collapse = F` is if you have data that has already been collapsed down to a standard reference population's age bins. **This is uncommon.** It isn't worth your time and energy to manually collapse the data -- so don't do it! This functionality is here just in case you receive data that has already been structured this way.

**Let's reuse the dataset with age, disease, pop, and gender from example #2**

```{r}
#| warning: false
#| message: false
temp4 <- copy(temp2)

```

**Collapse the data down to the same age bins (`agecat`) as those used in the default standard reference population**

```{r}
#| warning: false
#| message: false
#| eval: true
#| results: 'hide'
temp4[age == 0, agecat := "0"]
temp4[age %in% 1:4, agecat := "1-4 years"]
temp4[age %in% 5:14, agecat := "5-14 years"]
temp4[age %in% 15:24, agecat := "15-24 years"]
temp4[age %in% 25:34, agecat := "25-34 years"]
temp4[age %in% 35:44, agecat := "35-44 years"]
temp4[age %in% 45:54, agecat := "45-54 years"]
temp4[age %in% 55:64, agecat := "55-64 years"]
temp4[age %in% 65:74, agecat := "65-74 years"]
temp4[age %in% 75:84, agecat := "75-84 years"]
temp4[age %in% 85:100, agecat := "85+ years"]

temp4 <- temp4[, .(count = sum(disease), pop = sum(pop)), .(agecat, gender)]

head(temp4[])
```

```{r}
#| warning: false
#| message: false
#| eval: true
#| echo: false
pretty_kable(head(temp4[]))
```

**Now run age_standardize with `collapse = F`**

```{r}
#| warning: false
#| message: false
#| eval: true
#| results: 'hide'
ex4.1 <- age_standardize(ph.data = copy(temp4),
                         ref.popname = "2000 U.S. Std Population (11 age groups)",
                         collapse = F, # because already collapsed
                         my.count = "count", 
                         my.pop = "pop", 
                         per = 1000, 
                         conf.level = 0.95, 
                         by = "gender")

head(ex4.1)
```

```{r}
#| warning: false
#| message: false
#| eval: true
#| echo: false
pretty_kable(head(ex4.1))
```

Note the results in ex4.1 are exactly the same as those in ex2.2. This makes sense since age_standardize() collapsed the data in a similar (but more efficient manner) when creating ex2.2.

------------------------------------------------------------------------

## Example #5: Real world analysis

You're staffing data requests today and receive the following message: "*Hi! For a class, I'd like to get the 2019 15 to 18 birth rate for WA State. It would be appreciated if you could provide crude and age-standardized rates. If you could standardize to 'World (WHO 2000-2025) Std Million (single ages to 84)', that would be great! Thank you, S. Capstone*"

You quickly pull up [CHAT](https://www.doh.wa.gov/ForPublicHealthandHealthcareProviders/PublicHealthSystemResourcesandServices/CommunityHealthAssessmentandImprovement/CHAT) only to find that this specific indicator doesn't exist. In desperation, you resign yourself to using `rads`.

**Get birth counts for 15 to 18 year olds in 2019**

```{r}
#| warning: false
#| message: false
#| eval: true
#| results: 'hide'
  births <- rads.data::synthetic_birth[year == 2019 & age %in% 15:18, .(age)]
  
  births <- births[, .(births = .N), by = c("age")]  # collapse / aggregate
  
  setorder(births, age)
  
  head(births)
```

```{r}
#| warning: false
#| message: false
#| eval: true
#| echo: false
pretty_kable(head(births))
```

**Get the female populations corresponding to the birth counts**

```{r}
#| warning: false
#| message: false
#| eval: true
#| results: 'hide'
  pop <- rads.data::synthetic_population[gender == 'Female' &
                                         age %in% 15:18 & 
                                         year == 2019]  

  pop <- pop[, .(pop = sum(pop)), .(age)] # aggregate because it has rows for each race
  
  head(pop)
```

```{r}
#| warning: false
#| message: false
#| eval: true
#| echo: false
pretty_kable(head(pop))
```

**Merge population onto to birth counts**

```{r}
#| warning: false
#| message: false
#| echo: true
#| results: 'hide'
  temp5 <- merge(births, pop, by = "age", all = T)
  
  head(temp5)
```

```{r}
#| warning: false
#| message: false
#| eval: true
#| echo: false
pretty_kable(head(temp5))
```

**Run `age_standardize()`**

```{r}
#| warning: false
#| message: false
#| eval: true
#| results: 'hide'
  ex5.1 <- age_standardize(ph.data = temp5,
                           ref.popname = "World (WHO 2000-2025) Std Million (single ages to 84)", 
                           collapse = T,
                           my.count = "births", 
                           my.pop = "pop", 
                           per = 1000, 
                           conf.level = 0.95)
  head(ex5.1)
```

```{r}
#| echo: false
pretty_kable(head(ex5.1))
```

According to your analysis, the crude rate (`r rads::round2(ex5.1$crude.rate, 2)` / 1,000) is slightly higher than the adjusted rate (`r rads::round2(ex5.1$adj.rate, 2)` / 1,000).

------------------------------------------------------------------------

### (OPTIONAL) Calculating the standard error (SE) and relative standard error (RSE)

The method used by rads to calculate the confidence intervals is that recommended by WA DOH, i.e., the [Fay-Feuer method](https://wonder.cdc.gov/controller/pdf/FayFeuerConfidenceIntervals.pdf). Unfortunately, this method does not have a corresponding SE. We consulted with a WA DOH biostatistician who said, **"*SE's are not particularly useful for approximating the sampling distribution on the scale of adjusted rates*."** We therefore calculate **approximations** of the SE and RSE, as suggested by various state health departments. This means that the SE will be internally inconsistent with the confidence intervals ... *c'est la vie!*

For reference, here are the approximations: $$SE = {\text{adjusted rate} \over{\sqrt{{\text{number of cases}}}}}$$

$$RSE = {1 \over{\sqrt{\text{number of cases}}}}$$

------------------------------------------------------------------------

## Conclusion

If you've walked through this vignette and more or less understood what's going on, you're in good shape! If you're still confused, please walk through it again and then reach out if you still have questions. Good luck!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
