---
title: "BRFSS Functions"
format: gfm
prefer-html: true
self-contained: true
editor: visual
---

# Introduction

The Behavioral Risk Factor Surveillance System (BRFSS) is a gold mine of public health data -- but like any mine, you need the right tools to extract the value. Since BRFSS is a complex survey, analyses need to account for the survey design and weights to get accurate results. The survey design includes stratification and weighting to ensure the sample represents the full population, accounting for who was more or less likely to be included in the survey. When we want to analyze multiple years together (which we often do to increase precision), we need to adjust those survey weights to avoid overestimating our population.

This vignette will show you how to easily work with King County BRFSS data while properly handling all these survey design considerations. Don't worry - the functions do the heavy lifting for you! We'll cover everything from finding available variables to getting properly weighted estimates.

Note that the [BRFSS ETL process has its own repository](https://github.com/PHSKC-APDE/BRFSS) and questions regarding the data should be directed to the data steward.

# Load essential packages

```{r}
#| warning: false
#| message: false
library(rads)
library(data.table)
```

```{r echo=FALSE}
#| warning: false
#| message: false

# function to beautify tables
library(kableExtra)
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

# Checking Variable Availability Across Years

One quirk of BRFSS data is that not all questions are asked every year. Before diving into analysis, it's helpful to check which variables are available for your time period of interest. The `list_dataset_columns()` function makes this easy.

For example, let's see what variables are available for 2023:

```{r}
#| echo: true
#| output: false
# Check variables for a single year
vars_2023 <- list_dataset_columns("brfss", year = 2023)
head(vars_2023)
nrow(vars_2023)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(vars_2023))
nrow(vars_2023)
```

If you're planning to analyze multiple years together, you can check variable availability across years:

```{r}
#| echo: true
#| output: false
# Check variables across multiple years
vars_2019_2023 <- list_dataset_columns("brfss", year = 2019:2023)
head(vars_2019_2023)
nrow(vars_2019_2023)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(vars_2019_2023))
nrow(vars_2019_2023)
```

Notice that the `year(s)` column is not constant because BRFSS does not ask every question in every year.

# Getting BRFSS Data

There are two equivalent ways to get BRFSS data: using `get_data('brfss')` or `get_data_brfss()`. Both functions will:

1.  Load the data you request
2.  Automatically adjust weights if you're analyzing multiple years
3.  Survey-set the data so it's ready for analysis
4.  Return a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object, which is a data.table friendly survey object

Let's see both methods in action:

## Method 1: Using `get_data()`

This is the general interface that you can use to access any of APDE's analytic ready data.

```{r}
brfss_full <- get_data(
  dataset = "brfss",
  cols = c("chi_year", "age", "race4", "chi_sex", "prediab1"),
  year = 2019:2023
)
```

## Method 2: Using `get_data_brfss()` directly

```{r}
brfss_full_alt <- get_data_brfss(
  cols = c("chi_year", "age", "race4", "chi_sex", "prediab1"),
  year = 2019:2023
)
```

## Confirm that the two methods are identical

```{r}
if(identical(brfss_full, brfss_full_alt)){
  cat('\U0001f642 The two methods provide identical BRFSS data')
}
```

Both methods return an identical [`dtsurvey`](https://github.com/PHSKC-APDE/dtsurvey) object that's ready for analysis with [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc).

Notice that the functions provide an informative message regarding the survey object parameters. These will be hidden in the examples below, but are always produced when getting or survey setting BRFSS data.

Since BRFSS weights are designed to represent the population, we can verify that our multi-year weight adjustments are working properly by comparing population sizes. We expect the adjusted weights to represent an "average" population that falls between the earliest and latest years' populations since King County's population has been growing. Let's verify our weight adjustments are working as expected:

## Calculate the survey population at the beginning and end of the period

```{r}
pop_2019 <- sum(brfss_full[chi_year == 2019]$finalwt1)
pop_2023 <- sum(brfss_full[chi_year == 2023]$finalwt1)
```

## Calculate the adjusted population for the combined period

```{r}
pop_adjusted <- sum(brfss_full$default_wt)
```

## Is the value of the adjusted population between that for 2019 and 2023?

```{r}
if(pop_2023 > pop_adjusted & pop_adjusted > pop_2019){
  cat('\U0001f642 pop_adjusted is between pop_2019 and pop_2023')
}
```

# Working with HRAs and Regions

BRFSS data presents a unique challenge when analyzing Health Reporting Areas (HRAs) because it comes with ZIP codes rather than HRA assignments. Since ZIP codes don't perfectly align with HRA boundaries, we need to account for this uncertainty in our analyses.

To handle this, we use a statistical technique called multiple imputation. When you request HRA-related columns (`hra20_id`, `hra20_name`, or `chi_geo_region`), the function returns an [`imputationList`](https://cran.r-project.org/web/packages/mitools/mitools.pdf) object containing 10 different versions of the data. Each version represents a different possible way that ZIP codes could be assigned to HRAs based on their overlap. This approach allows us to capture the uncertainty in our geographic assignments and incorporate it into our statistical estimates.

Note: APDE decided to use 10 imputations based on an extensive empirical assessment by Daniel Casey. This is fixed in the ETL process and is not configurable.

```{r message=FALSE}
# Get data including HRA information
brfss_hra <- get_data_brfss(
  cols = c("chi_year", "age", "race4", "chi_sex", "prediab1", "obese", "hra20_name"),
  year = 2019:2023
)

# Confirm we generated an imputationList of 10 dtsurvey objects
if(inherits(brfss_hra, "imputationList") & 
   length(brfss_hra$imputations) == 10 & 
   inherits(brfss_hra$imputations[[1]], "dtsurvey")){
  cat('\U0001f642 brfss_hra is an imputationList of 10 dtsurvey objects')
}
```

Don't worry if this seems complex - the [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc) function automatically handles these [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf) objects.

# Modifying BRFSS Data

There are times when you might need to modify BRFSS data. For example, you might want to create a new variable. Before making any modifications, first consider whether your changes should be standardized. If you're creating variables that will be used across multiple projects (CHI, CHNA, Communities Count, etc.) or repeatedly year after year, contact the [BRFSS ETL steward](https://github.com/PHSKC-APDE/BRFSS) and politely request the addition of these changes to the analytic ready dataset.

For truly custom analyses, your modification approach will depend on whether you're working with a simple [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object or an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf). Let's look at each case:

## Modifying a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey)

You can modify a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object using data.table commands without disrupting its survey settings. If you want to use dplyr commands, you may break the internals of the [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) and would be wise to survey set it again following the instruction in the "Survey Setting and Creating Custom Weights" section below.

Regardless of whether you use data.table or dplyr commands, you are strongly encouraged to create new variables as needed rather than overwriting and deleting existing ones.

## Modifying an [ImputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf)

When working with HRA or region data, modifications become more complex since we need to maintain consistency across all 10 imputed datasets. Here's the step-by-step example that you can follow to help you in this process:

### 1. Get your BRFSS [ImputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf)

```{r message = FALSE}
brfss <- get_data_brfss(
  cols = c("age", "hra20_id"),
  year = 2019:2023
)
inherits(brfss, "imputationList") # confirms it is an imputationList
```

### 2. Keep the first imputed [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey)/data.table object

```{r}
brfss <- brfss$imputations[[1]]
inherits(brfss, 'dtsurvey')
inherits(brfss, 'data.table')
```

### 3. Create or modify a variable

In this step, the same guidelines apply that were mentioned in the 'Modifying a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey)' section above.

```{r}
# data.table::fcase is an implementation of SQL's CASE WHEN & comparable to dplyr::case_when
brfss[, age_category := fcase(age %in% 18:66, "working age",
                              age >= 67, "retirement age",
                              default = NA_character_
)]
```

### 4. Drop existing HRA or region variables (because they will be replaced)

```{r}
brfss[, intersect(c('hra20_id', 'hra20_name', 'chi_geo_region'), names(brfss)) := NULL]
```

### 5. Convert your [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object to a list of 10 dtsurvey objects

This step transforms your single [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object into a list of 10 [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) objects. Each object represents a different potential assignment of ZIP codes to HRAs. The [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object contains 10 columns of hra20_ids, and we create a separate table for each one. We merge on `hra20_name` and `chi_geo_region` to maintain geographic consistency between HRAs and regions.

```{r}
xwalk_hra_region <- rads.data::spatial_hra20_to_region20[, c("hra20_id", "hra20_name", "region_name")]

brfss <- lapply(1:10, function(i) {
  temp_dt <- copy(brfss)
  temp_dt[, hra20_id := get(paste0("hra20_id_", i))]
  temp_dt <- merge(
    temp_dt,
    xwalk_hra_region,
    by = "hra20_id",
    all.x = TRUE,
    all.y = FALSE
  )
  setnames(temp_dt, "region_name", "chi_geo_region")
  return(temp_dt)
})
```

### 6. Convert a standard list to an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf)

```{r}
brfss <- mitools::imputationList(brfss)
```

# Survey Setting and Creating Custom Weights

You might need to use `pool_brfss_weights()` in two scenarios:

1.  When analyzing specific years where certain questions were asked
2.  When you need to restore proper survey settings after using non-data.table commands for data manipulation

While `get_data_brfss()` automatically creates weights and survey sets imported data, you can create new weights and re-survey set the data using `pool_brfss_weights()`. Here are brief argument descriptions, see the `pool_brfss_weights()` help file for details:

-   `ph.data`: Your BRFSS dataset (can be a data.frame, data.table, [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey), or [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf))
-   `years`: Vector of years you want to analyze together
-   `year_var`: Name of the year column (defaults to 'chi_year')
-   `old_wt_var`: Name of the original weight variable (defaults to 'finalwt1')
-   `new_wt_var`: Name for your new weight variable
-   `wt_method`: Name of the method used to rescale your weights. Options include 'obs', 'pop', and 'simple' (defaults to 'obs')
-   `strata`: Name of the strata variable (defaults to 'x_ststr')

Let's see it in action:

```{r message=FALSE}
#| echo: true
#| output: false
# Create weights for odd years only
brfss_odd_years <- pool_brfss_weights(
  ph.data = brfss_full,
  years = c(2019, 2021),
  new_wt_var = "odd_year_wt"  # Name for the new weight variable
)

# Verify the adjusted population falls between the estimated single year populations
pop_2019 <- sum(brfss_odd_years[chi_year == 2019]$finalwt1)
pop_2021 <- sum(brfss_odd_years[chi_year == 2021]$finalwt1)
pop_2019_2021 <- sum(brfss_odd_years$odd_year_wt)
comparisons <- data.table(Description = c('2019', '2021', '2019, 2021'), 
                          `Pop 18-100` = c(pop_2019, pop_2021, pop_2019_2021))
comparisons[, `Pop 18-100` := format(`Pop 18-100`, big.mark = ',')]
head(comparisons)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(comparisons))
```

# Analyzing BRFSS Data with [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc)

Now for the fun part - analyzing our data! The [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc) function handles all the survey design considerations for us. Let's look at some examples:

## Calculate prediabetes prevalence by sex and race (using a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object)

```{r}
#| echo: true
#| output: false
prediab_by_group <- calc(
  ph.data = brfss_full,
  what = "prediab1",
  by = c("chi_sex", "race4"),
  metrics = c("mean", "rse"),
  proportion = TRUE  # Since prediab is binary
)
head(prediab_by_group)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(prediab_by_group))
```

## Calculate prediabetes prevalence by HRA20 (using an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf))

```{r}
#| echo: true
#| output: false
prediab_by_hra20 <- calc(
  ph.data = brfss_hra,
  what = "prediab1",
  by = c("hra20_name"),
  metrics = c("mean", "rse"),
  proportion = TRUE  
)
head(prediab_by_hra20)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(prediab_by_hra20[!is.na(hra20_name)]))
```

As noted in the [calc() wiki](https://github.com/PHSKC-APDE/rads/wiki/calc#example-analyses-with-resampled-datamultiple-imputation), when working with an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf), the proportion argument is ignored. However, we include it here to maintain consistent [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc) usage regardless of whether you're working with a [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) object or an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf).

## Calculate prediabetes & obesity prevalence by HRA20 & sex (using an [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf))

```{r}
#| echo: true
#| output: false
prediab_obese_hra20_sex <- calc(
  ph.data = brfss_hra,
  what = c("prediab1", "obese"),
  by = c("hra20_name", "chi_sex"),
  metrics = c("mean", "rse"),
  proportion = TRUE  
)
head(prediab_obese_hra20_sex)
```

```{r}
#| echo: false
#| output: true
pretty_kable(head(prediab_obese_hra20_sex[!is.na(hra20_name)]))
```

## Comparing `calc(..., where = ...)` with `pool_brfss_weights`

Those with experience using `calc()` might be wondering, "Why would we need to use `pool_brfss_weights()` to analyze a subset of years when we could just use the `where` argument in `calc`?" The short answer is that the methods are identical -- as long as you are only interested in the mean, standard error, RSE, and confidence intervals. However, if you want to know the survey weighted number of people within a given demographic or with a condition, you need to use `pool_brfss_weights()`. The following example analyzing data for 2022 compares the results from the two methods.

### Setting Up the Comparison

```{r}
#| echo: true
#| output: false
# Get 5 years of BRFSS data
brfss_where <- get_data_brfss(cols = c('chi_year', 'obese'), year = 2019:2023)

# Survey set a copy of the data for 2022
brfss_pooled <- pool_brfss_weights(ph.data = brfss_where, years = 2022, new_wt_var = 'wt_2022')
```

Use `calc()` to generate 2022 obesity prevalence

```{r}
#| echo: true
#| output: false
method_where <- calc(ph.data = brfss_where,
                    what = 'obese',
                    where = chi_year == 2022,
                    metrics = c("mean", "rse", "total"),
                    proportion = TRUE )

method_pooled <- calc(ph.data = brfss_pooled,
                    what = 'obese',
                    metrics = c("mean", "rse", "total"),
                    proportion = TRUE )

```

### Comparing Mean Estimates

Check if the mean, standard error, RSE, and CI values are equal

```{r}
#| echo: true
#| output: true
all.equal(method_where[, .(variable, mean, mean_se, mean_lower, mean_upper, rse)], 
          method_pooled[, .(variable, mean, mean_se, mean_lower, mean_upper, rse)])

```

### Comparing Population Totals

Check if the survey weighted 'total' values are equal

```{r}
#| echo: true
#| output: true
all.equal(method_where[, .(variable, total, total_se, total_lower, total_upper)], 
          method_pooled[, .(variable, total, total_se, total_lower, total_upper)])
```

### Key Takeaway

This shows us that the mean, standard error, RSE, and CI are identical for the two methods but the totals differ. Please remember, ***to get the correct survey weighted population you must use `pool_brfss_weights`***.

## Suppression & Reliability

Please refer to the [APDE_SmallNumberUpdate.xlsx](https://kc1.sharepoint.com/:x:/r/teams/DPH-APDEData/_layouts/15/Doc.aspx?sourcedoc=%7B44562E46-6E45-44B1-9BAC-38EED75E9222%7D&file=APDE_SmallNumberUpdate.xlsx&action=default&mobileredirect=true&DefaultItemOpen=1) file on SharePoint for details.

# Common Gotchas

1.  **Variable Availability**: Always check if your variables of interest are available in all years you want to analyze. Some BRFSS questions are only asked in certain years.

2.  **Survey Weights**: Make sure you're using the appropriate weights for your analysis period. If you subset your data to specific years after loading it, you will need to recalculate weights using `pool_brfss_weights()`.

3.  **Survey Design**: Always use [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc) for analysis to ensure proper handling of the survey design. Don't calculate means or totals directly from the data.

4.  **HRA / Region Analyses**: When working with HRAs or regions, you'll get an [`imputationList`](https://cran.r-project.org/web/packages/mitools/mitools.pdf). Don't worry -- [`calc()`](https://github.com/PHSKC-APDE/rads/wiki/calc) knows how to handle this! Just analyze it like you would any other BRFSS data.

# Conclusion

Working with BRFSS data requires careful attention to survey weights and design, but the functions we've covered make this process straightforward. Remember:

-   Check variable availability with `list_dataset_columns()`
-   Get data with `get_data_brfss()` or `get_data()`
-   Modify [dtsurvey](https://github.com/PHSKC-APDE/dtsurvey) or [imputationList](https://cran.r-project.org/web/packages/mitools/mitools.pdf) objects using data.table syntax
-   Create custom weights if needed with `pool_brfss_weights()`
-   Analyze using `calc()`

Happy analyzing!

-- *`r paste0("Updated by ", Sys.getenv("USERNAME"), ", ", Sys.Date())`*
