---
title: "Calculating rates with rads"
format:
  html:
    embed-resources: true
  gfm:
    prefer-html: false
vignette: >
  %\VignetteIndexEntry{Calculating rates with rads}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{knitr}
editor: source
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(rads)
library(data.table)
```

```{r}
#| echo: false
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

# Introduction

Many public health metrics are simple counts or means/proportions/fractions. The `rads::calc()` function was created to calculate these metrics from survey and line-level data. However, there are certain metrics and datasets (e.g., death & CHARS) where **true rates --** ***with a population denominator*** -- are of interest. Most of the time these rates need to be [age-adjusted](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) and most of the time we adjust to the US 2000 Standard Population with 11 age groups (see `rads.data::population_reference_pop_11_age_groups`). The [rads package](https://github.com/PHSKC-APDE/rads/) has a suite of tools that were developed to facilitate these types of analyses. This vignette explains how to use these tools in **four steps**.

# The four steps: an overview

**STEP 1:** Create a **summary table of event counts** (deaths, hospitalizations, etc.) **by single year age groups and strata** of interest. For example, part of your table might look like this:

| event_name | counts | age | year | region  |
|:-----------|:-------|:----|:-----|:--------|
| CVD death  | 26     | 47  | 2021 | Seattle |
| CVD death  | 17     | 47  | 2021 | North   |
| CVD death  | 28     | 47  | 2021 | East    |
| CVD death  | 36     | 47  | 2021 | South   |
| CVD death  | 23     | 48  | 2021 | Seattle |
| CVD death  | 28     | 48  | 2021 | North   |
| CVD death  | 36     | 48  | 2021 | East    |
| CVD death  | 24     | 48  | 2021 | South   |

**STEP 2:** Create a **table of population** denominators corresponding to the ages and demographics in the table created in Step 1.

| age | year | region  | population |
|:----|:-----|:--------|:-----------|
| 47  | 2021 | Seattle | 11353      |
| 47  | 2021 | North   | 347        |
| 47  | 2021 | East    | 8203       |
| 47  | 2021 | South   | 6432       |
| 48  | 2021 | Seattle | 3543       |
| 48  | 2021 | North   | 3041       |
| 48  | 2021 | East    | 425        |
| 48  | 2021 | South   | 7124       |

**STEP 3: Merge** the denominators (population) onto the table with the numerators (counts) using age and the strata of interest.

| event_name | counts | age | year | region  | population |
|:-----------|:-------|:----|:-----|:--------|:-----------|
| CVD death  | 26     | 47  | 2021 | Seattle | 11353      |
| CVD death  | 17     | 47  | 2021 | North   | 347        |
| CVD death  | 28     | 47  | 2021 | East    | 8203       |
| CVD death  | 36     | 47  | 2021 | South   | 6432       |
| CVD death  | 23     | 48  | 2021 | Seattle | 3543       |
| CVD death  | 28     | 48  | 2021 | North   | 3041       |
| CVD death  | 36     | 48  | 2021 | East    | 425        |
| CVD death  | 24     | 48  | 2021 | South   | 7124       |

**STEP 4:** Use [**rads::age_standardize()**](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) to calculate the crude and age-adjusted rates

# The four steps: example 1

As a simple example, let's assess the King County death rate from falls in 2019-2023, by intent.

### STEP 1: Create a summary **table of event counts**

We will load sample death data using `rads.data::synthetic_death` and use `rads::death_injury_matrix_count` to process the ICD-10 death codes into human readable mechanisms and intents.

Let's grab the data from [`rads.data`](https://github.com/PHSKC-APDE/rads.data):

```{r}
deaths <- rads.data::synthetic_death

deaths <- deaths[year %in% 2019:2023]

deaths <- deaths[, .(year, age, underlying_cod_code)]
```

Now, let's identify the deaths from falls using the `rads::death_injury_matrix_count` function. To learn about this function, type `?death_injury_matrix_count` in your console or walk through the [death_functions wiki](https://github.com/PHSKC-APDE/rads/wiki). For now, just trust that this is the correct syntax:

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
deaths <- death_injury_matrix_count(ph.data = deaths,
                                    intent = "*", # all five intents
                                    mechanism = "fall",
                                    icdcol = 'underlying_cod_code',
                                    by = 'age')
head(deaths[deaths > 9])
```

```{r}
#| echo: false
pretty_kable(head(deaths[deaths > 9]))
```

Glancing at the output from the `death_injury_matrix_count` function, we see that we have **aggregated** `deaths` by `age` and stratified by `intent`. This is all we really need. Let's tidy our data set and rename the `deaths` column:

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
deaths <- deaths[, .(mechanism, intent, age, count = deaths)]
head(deaths[count > 9])
```

```{r}
#| echo: false
pretty_kable(head(deaths[count > 9]))
```

Congratulations, you've completed step 1.

### STEP 2: Create a table of **population denominators**

Let's load some population data using `rads.data::synthetic_pop`

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
population <- rads.data::synthetic_population

population <- population[year %in% 2019:2023]
  
head(population)
```

```{r}
#| echo: false
pretty_kable(head(population))
```

Note that the populations are for individual years, but we want the population for the entirety of 2019-2023. We also only need the `age` and `pop` columns for our analysis. So, let's aggregate the `pop` by `age`. 
```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
population <- population[, .(pop = sum(pop)), .(age)]
head(population)
```

```{r}
#| echo: false
pretty_kable(head(population))
```

Since we are interested in strata (i.e., `intent`), we will need a complete set of population data for each stratum.

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
population <- rbindlist(lapply(X = 1:length(unique(deaths$intent)),
                               FUN = function(X){
                                 temp <- copy(population)[, intent := unique(deaths$intent)[X]]
                                 }))
population[, .N, intent] # confirm that have 101 rows per intent
```

```{r}
#| echo: false
pretty_kable(population[, .N, intent])
```

### STEP 3: **Merge** the denominators onto the numerators

The title says it all:

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
deaths <- merge(deaths, 
                population, 
                by = c('age', 'intent'), 
                all.x = F, # drop if death strata do not match a population
                all.y = T) # keep population data if do not have deaths
deaths[is.na(count), count := 0] # formally set rows with zero counts to zero
head(deaths[count > 9]) # only display non-suppressed data
```

```{r}
#| echo: false
pretty_kable(head(deaths[count > 9]))
```

### STEP 4: Use [rads::age_standardize()](https://github.com/PHSKC-APDE/rads/wiki/age_standardize)

The documentation tells us that `rads::age_standardize` needs a data set with either an `age` or `agecat` column. Since we have the former, we are good to go as long as we choose the correct reference population. Typing `list_ref_pop` will show you all the available reference populations, however we already know that DOH uses the 2000 US standard population with 11 age groups, so we can plug that into the function.

```{r}
#| message: false
#| warning: false
#| echo: true
#| results: 'hide'
est <- age_standardize(ph.data = deaths,
                       ref.popname = "2000 U.S. Std Population (11 age groups)",
                       collapse = T,
                       my.count = 'count',
                       my.pop = 'pop',
                       per = 100000,
                       conf.level = 0.95, 
                       by = 'intent')
head(est)
```

```{r}
#| echo: false
pretty_kable(head(est))
```

Here we see that the King County **crude** and **adjusted** fall death rates are relatively similar. We also see that the vast majority of fall related deaths are unintentional. 

# The four steps: example 2

In this example we will calculate the adjusted rate of deaths from ***any cause*** by King County in 2019, stratified by race. 

### STEP 1: Create a summary **table of event counts**

```{r}
#| message: false
#| warning: false
deaths <- rads.data::synthetic_death
deaths <- deaths[year == 2019]
deaths <- deaths[, .(year, age, race)]
head(deaths)
```

```{r}
#| echo: false
pretty_kable(head(deaths))
```

Remember that step 1 is supposed to create a **summary** table. So, let's collapse/aggregate/sum the data to get the count of deaths by age and race:

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
deaths <- deaths[, .(count = .N), .(age, race)]
head(deaths)
```

```{r}
#| echo: false
pretty_kable(head(deaths))
```

Finally, let's take a peek at the number of deaths by race:

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
deaths[, .(count = sum(count)), race]
```

```{r}
#| echo: false
pretty_kable(deaths[, .(count = sum(count)), race])
```

This table of deaths by race shows that the number of deaths among White residents (`r format(sum(deaths[race == 'White']$count), big.mark = ',')`) is more than 18 times higher than that among the next highest racial group (Asian = `r sum(deaths[race == 'Asian']$count)`). Does that mean the death *rates* among White residents are also exceptionally high? Let's get the corresponding population data and find out.

### STEP 2: Create a table of **population denominators**

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
population <- rads.data::synthetic_population

population <- population[year == 2019]

# aggregate to pop per race and age
population <- population[, .(pop = sum(pop)), .(race, age)]

head(population)
```

```{r}
#| echo: false
pretty_kable(head(population))
```

Let's confirm that we have complete population data set (ages 0 to 100) for each race.

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
population[, .N, race]
```

```{r}
#| echo: false
pretty_kable(population[, .N, race])
```

### STEP 3: **Merge** the denominators onto the numerators

Just do it!

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
deaths <- merge(deaths, 
                population, 
                by = c('race', 'age'), 
                all.x = F, 
                all.y = T) 
deaths[is.na(count), count := 0] # set rows with zero counts to zero
head(deaths)
```

```{r}
#| echo: false
pretty_kable(head(deaths))
```

### STEP 4: Use [rads::age_standardize()](https://github.com/PHSKC-APDE/rads/wiki/age_standardize)

```{r}
#| message: false
#| warning: false
#| paged.print: false
#| echo: true
#| results: 'hide'
est <- age_standardize(ph.data = deaths,
                       ref.popname = "2000 U.S. Std Population (11 age groups)",
                       collapse = T,
                       my.count = 'count',
                       my.pop = 'pop',
                       per = 100000,
                       conf.level = 0.95, 
                       by = 'race')
head(est)
```

```{r}
#| echo: false
pretty_kable(head(est))
```

Reviewing the age adjusted rates (`adj.rate`) in this *synthetic data analysis*, we see that the rate of death for White residents is middle of the road (`r format(round(est[race == 'White']$adj.rate), big.mark = ',')`). Their rate is much higher than those for Asian residents (`r format(round(est[race == 'Asian']$adj.rate), big.mark = ',')`) and lower than those for American Indian Alaska Native residents (`r format(round(est[race == 'AIAN']$adj.rate), big.mark = ',')`).

### (OPTIONAL) Calculating the standard error (SE) and relative standard error (RSE)

The method used by rads to calculate the confidence intervals is that recommended by WA DOH, i.e., the [Fay-Feuer method](https://wonder.cdc.gov/controller/pdf/FayFeuerConfidenceIntervals.pdf). Unfortunately, this method does not have a corresponding SE. APDE consulted with a WA DOH biostatistician who said, **"*SE's are not particularly useful for approximating the sampling distribution on the scale of adjusted rates*."** We therefore calculate **approximations** of the SE and RSE, as suggested by various state health departments. This means that the SE will be internally inconsistent with the confidence intervals ... *c'est la vie!*

For reference, here are the approximations:
$$SE = {\text{adjusted rate} \over{\sqrt{{\text{number of cases}}}}}$$

$$RSE = {1 \over{\sqrt{\text{number of cases}}}}$$

# Conclusion

We hope this vignette has convinced you of two things:

1.  Age-adjusted rates are important because they can (sometimes) tell a different story compared to crude rates and counts!
2.  Calculating age-adjusted rates within the PHSKC universe is readily doable if you follow the four basic steps:
    1.  Create a summary table by age and strata of interest

    2.  Create an appropriate matching population table

    3.  Merge the population data onto the summary table

    4.  Use `rads::age_standardize()`

If you've walked through this vignette and more or less understood what's going on, you're in good shape! If you're still confused, please walk through it again and then reach out if you still have questions. Good luck!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
