---
title: "Calculating rates with rads"
output:
  github_document: default
vignette: |
  %\VignetteEngine{knitr::knitr} %\VignetteIndexEntry{calculating_rates_with_rads}
---

```{r, echo = F, message=FALSE, warning=FALSE}
library(rads)
library(data.table)
```

```{r, echo=FALSE}
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

# Introduction

Many PHSKC metrics are simple counts or means/proportions/fractions. The `rads::calc()` function was created to calculate these metrics from survey and line-level data. However, there are certain metrics and datasets (e.g., death & CHARS) where **true rates --** **with a population denominator** -- are of interest. Most of the time these rates need to be [age-adjusted](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) and most of the time DOH and APDE adjust to the US 2000 Standard Population with 11 age groups (see `rads.data::population_reference_pop_11_age_groups`). The [rads package](https://github.com/PHSKC-APDE/rads/) has a suite of tools that were developed to facilitate these types of analyses. This vignette explains how to use these tools in **four steps**.

# The four steps: an overview

**STEP 1:** Create a **summary table of event counts** (deaths, hospitalizations, etc.) **by single year age groups and strata** of interest. For example, part of your your table might look like this:

| event_name | counts | age | year | region  |
|:-----------|:-------|:----|:-----|:--------|
| CVD death  | 26     | 47  | 2021 | Seattle |
| CVD death  | 17     | 47  | 2021 | North   |
| CVD death  | 28     | 47  | 2021 | East    |
| CVD death  | 36     | 47  | 2021 | South   |
| CVD death  | 23     | 48  | 2021 | Seattle |
| CVD death  | 28     | 48  | 2021 | North   |
| CVD death  | 36     | 48  | 2021 | East    |
| CVD death  | 24     | 48  | 2021 | South   |

**STEP 2:** Create a **table of population** denominators corresponding to the ages and demographics in the table created in Step 1.

| age | year | region  | population |
|:----|:-----|:--------|:-----------|
| 47  | 2021 | Seattle | 11353      |
| 47  | 2021 | North   | 347        |
| 47  | 2021 | East    | 8203       |
| 47  | 2021 | South   | 6432       |
| 48  | 2021 | Seattle | 3543       |
| 48  | 2021 | North   | 3041       |
| 48  | 2021 | East    | 425        |
| 48  | 2021 | South   | 7124       |

**STEP 3: Merge** the denominators (population) onto the table with the numerators (counts) using age and the strata of interest.

| event_name | counts | age | year | region  | population |
|:-----------|:-------|:----|:-----|:--------|:-----------|
| CVD death  | 26     | 47  | 2021 | Seattle | 11353      |
| CVD death  | 17     | 47  | 2021 | North   | 347        |
| CVD death  | 28     | 47  | 2021 | East    | 8203       |
| CVD death  | 36     | 47  | 2021 | South   | 6432       |
| CVD death  | 23     | 48  | 2021 | Seattle | 3543       |
| CVD death  | 28     | 48  | 2021 | North   | 3041       |
| CVD death  | 36     | 48  | 2021 | East    | 425        |
| CVD death  | 24     | 48  | 2021 | South   | 7124       |

**STEP 4:** Use [**rads::age_standardize()**](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) to calculate the crude and age-adjusted rates

# The four steps: example 1

As a simple example, let's assess the King County death rate from falls in 2016 through 2020, by intent.

### STEP 1: Create a summary **table of event counts**

We will pull down the relevant death data using `rads::get_data_death` and use `rads::death_injury_matrix_count` to process the ICD-10 death codes into human readable mechanisms and intents.

Let's use `rads` to get our death data and check if it seems reasonable:

```{r}
deaths <- get_data_death(cols = c('chi_age', 'chi_year', 'nonseattle', 
                                  'underlying_cod_code'), 
                               year = c(2016:2020), 
                               kingco = T)
names(deaths) 
nrow(deaths) # should be ~65,000 based on apriori knowledge
```

We see that the column names and the number of rows are what we expected. Now, let's identify the deaths from falls using the `rads::death_injury_matrix_count` function. To learn about this function, type `?death_injury_matrix_count` in your console or walk through the [death_functions wiki](https://github.com/PHSKC-APDE/rads/wiki). For now, just trust that this is the correct syntax:

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
deaths <- death_injury_matrix_count(ph.data = deaths,
                               intent = "*", # all five intents
                               mechanism = "fall",
                               icdcol = 'underlying_cod_code',
                               kingco = F, # False because already subset
                               group_by = 'chi_age') 
head(deaths[deaths >9]) # only display non-suppressed data
```

```{r, echo=FALSE}
pretty_kable(head(deaths[deaths >9]))
```

Glancing at the output from the `death_injury_matrix_count` function, we see the that we have **aggregated** `deaths` by `chi_age` and stratified by `intent`. This is all we really need. Let's tidy our data set and rename the two essential columns at the same time:

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
deaths <- deaths[, .(intent, age = chi_age, count = deaths)]
head(deaths[count >9]) # only display non-suppressed data
```

```{r, echo=FALSE}
pretty_kable(head(deaths[count >9]))
```

The table above confirms shows that we created a summary table of event (fall deaths) counts by age and intent. Congratulations, you've completed step 1.

### STEP 2: Create a table of **population denominators**

We can easily get the corresponding population data using [rads::get_population()](https://github.com/PHSKC-APDE/rads/wiki/get_population)

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
population <- rads::get_population(years = c(2016:2020), 
                                   geo_type = 'kc', 
                                   group_by = c('ages'))
head(population)
```

```{r, echo=FALSE}
pretty_kable(head(population))
```

The essential columns of interest are `pop` and `age`, so let's drop the other columns.

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
population <- population[, .(age, pop)]
head(population)
```

```{r, echo=FALSE}
pretty_kable(head(population))
```

Since we are interested in strata (i.e., `intent`), we will need a complete set of population data for each stratum.

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
population <- rbindlist(lapply(X = 1:length(unique(deaths$intent)),
                               FUN = function(X){
                                 temp <- copy(population)[, intent := unique(deaths$intent)[X]]
                                 }))
population[, .N, intent] # confirm that have 101 rows per intent
```

```{r, echo=FALSE}
pretty_kable(population[, .N, intent])
```

### STEP 3: **Merge** the denominators onto the numerators

The title says it all:

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
deaths <- merge(deaths, 
                population, 
                by = c('age', 'intent'), 
                all.x = F, # drop if death strata do not match a population
                all.y = T) # keep population data if do not have deaths
deaths[is.na(count), count := 0] # formally set rows with zero counts to zero
head(deaths[count > 9]) # only display non-suppressed data
```

```{r, echo=FALSE}
pretty_kable(head(deaths[count > 9]))
```

### STEP 4: Use [rads::age_standardize()](https://github.com/PHSKC-APDE/rads/wiki/age_standardize)

The documentation tells us that `rads::age_standardize` needs a data set with either an `age` or `agecat` column. Since we have the former, we are good to go as long as we choose the correct reference population. Typing `list_ref_pop` will show you all the available reference populations, however we already know that that DOH uses the 2000 US standard population with 11 age groups, so we can plug that into the function.

```{r message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
est <- age_standardize(ph.data = deaths,
                       ref.popname = "2000 U.S. Std Population (11 age groups)",
                       collapse = T,
                       my.count = 'count',
                       my.pop = 'pop',
                       per = 100000,
                       conf.level = 0.95, 
                       group_by = 'intent')
head(est)
```

```{r, echo=FALSE}
pretty_kable(head(est))
```

Here we see that the King County **crude** and **adjusted** fall death rates can differ substantially (e.g., Unintentional) or slightly (e.g., Undetermined). 

# The four steps: example 2

In this example we will calculate the adjusted rate of deaths from any cause by King County Regions in 2019. Because Regions are not baked into the death data, we will have to crosswalk from HRAs to Regions.

### STEP 1: Create a summary **table of event counts**

```{r message=FALSE, warning=FALSE}
deaths <- get_data_death(cols = c('chi_age', 
                                  'chi_year',
                                  'apde_geo_hra2010_short', 
                                  'underlying_cod_code'), 
                               year = c(2019), 
                               kingco = T)
```

Now let's crosswalk from King County HRAs to Regions using `rads::get_walk()`.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
xwalk <- rads::get_xwalk(geo1 = 'region10', geo2 = 'hra10')

deaths <- merge(deaths, xwalk, 
                by.x = "apde_geo_hra2010_short", by.y = "hra10", 
                all.x = T, all.y = F)

deaths <- deaths[, .(age = chi_age, underlying_cod_code, Region = region10)]

nrow(deaths[is.na(Region)]) # rows with missing Region information
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
head(deaths)
```

```{r, echo=FALSE}
pretty_kable(head(deaths))
```

As you can see, there are few rows where the death data are missing a Region (due to a missing HRA). For simplicity sake, let's drop these rows.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
deaths <- deaths[!is.na(Region)]

```

Remember that step 1 is supposed to create a **summary** table. So, let's collapse/aggregate/sum the data to get the count of deaths by age and region:

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
deaths <- deaths[, .(count = .N), .(age, Region)]
head(deaths)
```

```{r, echo=FALSE}
pretty_kable(head(deaths))
```

Finally, let's take a peak at the number of deaths by region:

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
deaths[, sum(count), Region]
```

```{r, echo=FALSE}
pretty_kable(deaths[, sum(count), Region])
```

This table of deaths by region shows that the number of deaths in South King County is five times higher than those in North King County (`r deaths[Region == 'South', .(total = sum(count))][['total']]` vs. `r deaths[Region == 'North', .(total = sum(count))][['total']]`). Does that mean the death *rates* in South King County are five times higher than those in North King County? Let's get the corresponding population data and find out.

### STEP 2: Create a table of **population denominators**

We again use [rads::get_population()](https://github.com/PHSKC-APDE/rads/wiki/get_population) to the necessary population data

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
population <- rads::get_population(years = c(2019), 
                                   geo_type = 'region', 
                                   group_by = c('ages', 'geo_id'))
head(population)
```

```{r, echo=FALSE}
pretty_kable(head(population))
```

The key columns of interest are `pop`, `geo_id` (which has our regions), and `age`. The other columns are all constants, so let's just keep what we need.

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
population <- population[, .(Region = geo_id, age, pop)]
head(population)
```

```{r, echo=FALSE}
pretty_kable(head(population))
```

Note that, unlike in the example above, we already have a complete population data set (ages 0 to 100) for each strata.

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
population[, .N, Region]
```

```{r, echo=FALSE}
pretty_kable(population[, .N, Region])
```

### STEP 3: **Merge** the denominators onto the numerators

Just do it!

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
deaths <- merge(deaths, 
                population, 
                by = c('Region', 'age'), 
                all.x = F, 
                all.y = T) 
deaths[is.na(count), count := 0] # set rows with zero counts to zero
head(deaths)
```

```{r, echo=FALSE}
pretty_kable(head(deaths))
```

### STEP 4: Use [rads::age_standardize()](https://github.com/PHSKC-APDE/rads/wiki/age_standardize)

```{r message=FALSE, warning=FALSE, paged.print=FALSE, echo=TRUE, results='hide'}
est <- age_standardize(ph.data = deaths,
                       ref.popname = "2000 U.S. Std Population (11 age groups)",
                       collapse = T,
                       my.count = 'count',
                       my.pop = 'pop',
                       per = 100000,
                       conf.level = 0.95, 
                       group_by = 'Region')
head(est)
```

```{r, echo=FALSE}
pretty_kable(head(est))
```

Here we see that the the **crude rate is lower** in South King County (`r est[Region=="South"]$crude.rate`) compared to North King County (`r est[Region=="North"]$crude.rate`). This flips the relationship we saw with the counts (`r est[Region=="South"]$count` vs `r est[Region=="North"]$count`), but does not account for differences in the population structure in the two regions. In contrast, the **adjusted rate is higher** in South King County (`r est[Region=="South"]$adj.rate`) vs. North King County(`r est[Region=="North"]$adj.rate`) .

### (OPTIONAL) Calculating the standard error (SE) and relative standard error (RSE)

The method used by rads to calculate the confidence intervals is that recommended by WA DOH, i.e., the [Fay-Feuer method](https://wonder.cdc.gov/controller/pdf/FayFeuerConfidenceIntervals.pdf). Unfortunately, this method does not have a corresponding SE. APDE consulted with a WA DOH biostatistician who said, **"*SE’s are not particularly useful for approximating the sampling distribution on the scale of adjusted rates*."** We therefore calculate **approximations** of the SE and RSE, as suggested by various state health departments. This means that the SE will be internally inconsistent with the confidence intervals ... *c'est la vie!*

For reference, here are the approximations:
$$SE = {\text{adjusted rate} \over{\sqrt{{\text{number of cases}}}}}$$

$$RSE = {1 \over{\sqrt{\text{number of cases}}}}$$

# Conclusion

We hope this vignette has convinced you of two things:

1.  Age-adjusted rates are important because they can (sometimes) tell a different story compared to crude rates and counts!
2.  Calculating age-adjusted rates within the PHSKC universe is readily doable if you follow the four basic steps:
    1.  Create a summary table by age and strata of interest

    2.  Create an appropriate matching population table

    3.  Merge the population data onto the summary table

    4.  Use `rads::age_standardize()`

If you've walked through this vignette and more or less understood what's going on, you're in good shape! If you're still confused, please walk through it again and then reach out if you still have questions. Good luck!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
