---
title: "CHARS Functions"
format: gfm
prefer-html: false
self-contained: true
editor: visual
---

# Introduction

The [rads package](https://github.com/PHSKC-APDE/rads/) has a suite of tools designed to facilitate and accelerate the analysis of standardized CHARS (Comprehensive Hospital Abstract Reporting System) data. Combining the `rads` functions below with the clean CHARS data (or with practice data from `rads.data`, which is installed automatically with rads) should allow analysts to conduct custom analyses with relative ease. The core `rads` CHARS functions are:

-   `chars_icd_ccs()`: view available CHARS ICD-9-CM and ICD-10-CM descriptions as well as 'superlevel', 'broad', 'midlevel', and 'detailed' aggregations derived from [AHRQ's HCUP CCSR](https://hcup-us.ahrq.gov/toolssoftware/ccsr/ccs_refined.jsp) that can be used with `chars_icd_ccs_count()`
-   `chars_icd_ccs_count()`: generate counts of CHARS hospitalizations using ICD-9-CM or ICD-10-CM descriptions or 'superlevel', 'broad', 'midlevel', and 'detailed' categories.
-   `chars_injury_matrix()`: view all available intents and mechanisms that can be used with `chars_injury_matrix_count()` (2012+)
-   `chars_injury_matrix_count()`: generate counts of injury related hospitalizations by intent and mechanism (2012+)
-   `chars_validate_data()`: validate that your CHARS dataset has the proper structure and columns needed for the injury and ICD-CM analysis functions

Additionally, if you have APDE credentials and are working within King County's infrastructure, you can easily download standardized CHARS data from SQL into R (2012+) using `apde.data::chars()`.

All of these functions have detailed help files that are accessible by typing `?function_name`, e.g. `?chars_injury_matrix_count`. Some examples for how to use these functions are given below.

***A few quick notes before we begin ...***

-   `apde.data::chars()` can provide you with ICD-9-CM data (2012-2015) as well as ICD-10-CM data (2016+).
-   `chars_injury_matrix()` and `chars_injury_matrix_count()` are agnostic as to whether the underlying data are ICD-9-CM or ICD-10-CM.
-   `chars_icd_ccs()` & `chars_icd_ccs_count()` need you to specify which ICD-CM version you have in your data. This means you can analyze 2012-2015 data or 2016+ data, but not both at the same time in a single command.
-   If you want to create age-adjusted rates, we recommend you read the [age_standardize](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) and [calculating_rates_with_rads](https://github.com/PHSKC-APDE/rads/wiki/calculating_rates_with_rads) vignettes after working through this one.

# Set up the environment

```{r}
#| warning: false
#| message: false
rm(list=ls())
library(rads)
library(data.table)
```

```{r}
#| echo: FALSE
library(kableExtra)
```

# Getting CHARS data

## For practice and development: `rads.data::synthetic_chars`

If you're learning to work with CHARS data or developing new analysis code, we provide a synthetic dataset that mimics the structure of real CHARS data. This privacy-safe dataset is automatically available when you install `rads` and contains over 250,000 rows of injury-related diagnosis data.

```{r}
#| warning: false
#| message: false
# Load the synthetic CHARS data
data(synthetic_chars, package = "rads.data")

# View the structure
str(synthetic_chars)
```

The synthetic dataset includes the key columns needed for injury and ICD-CM analyses:

-   `seq_no`: unique patient-visit identifier
-   `diag1`: ICD-10-CM primary diagnosis code
-   `injury_nature_broad` & `injury_nature_narrow`: designation of the type of injury definition used. 'narrow' follows the CDC recommendation and 'broad' follows a more expansive definition. see `?rads.data::synthetic_chars` for details
-   `injury_intent`: injury intent classification (e.g., 'assault', 'unintentional')
-   `injury_mechanism`: injury mechanism classification (e.g., 'fall', 'firearm')
-   `chi_geo_kc`: King County indicator
-   `temperament`: a categorical 'demographic' indicator for practice stratification

This synthetic dataset is perfect for testing code before running analyses on real data.

## For APDE analysts: `apde.data::chars()`

***Note: This section is only relevant for analysts working within King County's APDE infrastructure who have the necessary credentials.***

`apde.data::chars()` takes nine potential arguments:

-   `cols`: character. the names of the columns that you want to download. Limiting this list to only the variables you truly need can significantly improve download speed. In most cases, this includes any demographic stratifiers of interest along with "seq_no", "diag1", "injury_nature_broad", "injury_nature_narrow", "injury_intent", and "injury_mechanism".
-   `year`: the year(s) of interest, from 2012 to the present.
-   `kingco`: logical (T\|F) . True or false specifies whether to limit the download to King County, based on **truncated** ZIP codes (980## and 981##).
-   `version`: character. Either `'final'` or `'stage'.`
-   `wastate`: logical (T\|F). When false, data will include Oregon.
-   `inpatient`: logical (T\|F). When false, data will include observation patients (i.e., outpatients).
-   `deaths`: logical (T\|F). When true, the data will include those who died while in the hospital.
-   `topcode`: logical (T\|F). When true, `chi_age` will be top coded to 100 to match population data top coding.

If you do not specify any of the arguments, you will get all CHARS data columns, for the latest year, for King County (defined by truncated ZIP codes), limited to inpatients, including those who died while hospitalized, with ages top coded to 100.

```{r}
#| warning: false
#| message: false
charsDT <- apde.data::chars(cols = c("seq_no", "diag1", "injury_nature_broad", 
                                     "injury_nature_narrow", "injury_intent", 
                                     "injury_mechanism", "chi_geo_kc", "chi_year", 
                                     "chi_age"), year = 2023)
unique(charsDT$chi_geo_kc) # confirm data is limited to King County
unique(charsDT$chi_year) # check the year
max(charsDT$chi_age, na.rm = T) # check top coding
```

For the remainder of this vignette, we'll use the synthetic dataset so that everyone can follow along:

```{r}
#| warning: false
#| message: false
charsDT <- rads.data::synthetic_chars
```

## :warning: A note about King County population denominators

**Important note:** Since King County in CHARS data is defined by truncated ZIP codes (980## and 981##), the correct denominator when calculating rates should be defined the same way. To be clear, this means you should **NOT** **use `apde.data::population(kingco = T)`** for King County CHARS denominators. Instead, you should get ZIP code population data and aggregate it for King County.

Here's an example of how to obtain a King County population denominator by age, gender, and race (with Hispanic as a race):

```{r}
#| warning: false
#| message: false
#| eval: false
# Get ZIP code level population data
denominator <- apde.data::population(kingco = FALSE, 
                             geo_type = 'zip', 
                             group_by = c('ages', 'genders', 'race_eth'))

# Subset to ZIP that begin with 980/981
denominator <- denominator[grepl('^980|^981', geo_id)]

# Sum the population across all these ZIP codes by gender, race/eth, and age
denominator <- denominator[, .(pop = sum(pop)), .(gender, race_eth, age)]

# Label it as King County
denominator[, geo_id := 'King County']
```

# chars_validate_data()

Before diving into analysis, it's good practice to validate that your CHARS dataset has the proper structure. The `chars_validate_data()` function checks that your data contains all the required columns with appropriate data types and values.

```{r}
#| warning: false
#| message: false
validated_chars <- chars_validate_data(ph.data = charsDT, 
                                       icdcol = 'diag1',
                                       icdcm_version = 10)
```

This function validates:

-   Required columns exist: `seq_no`, `injury_nature_broad`, `injury_nature_narrow`, `injury_intent`, `injury_mechanism`, and the ICD column
-   `seq_no` contains unique values (one per patient-visit)
-   Injury columns have appropriate data types (logical for nature columns, character for intent/mechanism)
-   ICD codes are properly formatted
-   Standard injury intent and mechanism categories are present (with informative messages if any are missing)

The function returns the validated data (with any necessary cleaning applied to ICD codes) and is especially useful when working with custom or external CHARS datasets.

# chars_icd_ccs()

`chars_icd_ccs()` takes three arguments:

-   `ref_typ`: specifies the hospital diagnosis descriptions that are of interest to you. Acceptable options include: 'all', 'icdcm', 'superlevel', 'broad', 'midlevel', & 'detailed'.
-   `icdcm_version`: specifies the ICD-CM version that you want to reference. Acceptable options include: 9 & 10, with 10 being the default.

Do not attempt to manually browse through `chars_icd_ccs()` ... you will lose your mind because it has more than 100,000 rows! Rather, use it to identify the type of non-injury hospitalization of interest. The structure is simple and (hopefully!) self-explanatory. Let's take a look at the first three rows as an example by typing `chars_icd_ccs()[1:3]`:

```{r}
#| echo: false
  kable(chars_icd_ccs()[1:3], format = "markdown") 
```

Teaching about [regular expression, a.k.a. *regex*](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html) and filtering is outside the bounds of this vignette. However, I imagine you will usually want to use aggregated hospitalization data so I encourage you to look at the unique values of superlevel, broad, midlevel, and detailed data. For example, let's examine the CCSR broad categories with `chars_icd_ccs(ref_type = 'broad')`:

```{r}
#| echo: false
  kable(chars_icd_ccs(ref_type = 'broad'), format = 'markdown')
```

# chars_icd_ccs_count()

`chars_icd_ccs_count()` allows the user to get CHARS counts by ICD-CM code, ICD-cm description, or the superlevel, broad, midlevel, and detailed categories. I provide examples of each of these below, in order of decreasing granularity / specificity using hypertensive heart disease as a case study.

```{r}
#| echo: FALSE
kable(chars_icd_ccs()[grep('I110', icdcm_code)], format = "markdown")
```

However, before we begin, let's review the possible arguments used by `chars_icd_ccs_count()`:

-   `ph.data`: the name of a person level data.table/data.frame of CHARS data with ICD10-cm codes
-   `icdcm_version`: specifies the ICD-CM version that you want to reference. Acceptable options include: 9 & 10, with 10 being the default.
-   `icdcm`: the ICD-CM code of interest OR its description. It is case insensivitive and partial strings are allowed.
-   `superlevel`: 'superlevel' level descriptions that are of interest. Case insensivitive and partial strings are allowed.
-   `broad`: CCSR derived 'broad' level descriptions that are of interest. Case insensivitive and partial strings are allowed.
-   `midlevel`: 'midlevel level descriptions that are of interest. Case insensivitive and partial strings are allowed.
-   `detailed`: CCSR derived 'detailed' level descriptions that are of interest. Case insensivitive and partial strings are allowed.
-   `icdcol`: the name of the column in `ph.data` that contains the ICD10-cm codes. Default is `diag1`, which is provided when you use `apde.data::chars()`.
-   `group_by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `kingco`: logical (T\|F) specifying whether to limit the data analysis to King County. Only works if ph.data still has the `chi_geo_kc` column.

## getting CHARS counts by ICD-10-CM code

```{r}
  mycode <- chars_icd_ccs_count(ph.data = charsDT, 
                                icdcm = 'I110')
```

```{r}
#| echo: FALSE
  mycode[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mycode, format = "markdown") 
```

## getting CHARS counts by ICD-10-CM description

```{r}
  mydesc <- chars_icd_ccs_count(ph.data = charsDT, 
                                icdcm = 'Hypertensive heart disease with heart failure')
```

```{r}
#| echo: FALSE
  mydesc[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mydesc, format = "markdown") 
```

These results are identical to searching by code because, for hypertensive heart failure, there is a one-to-one match of description to ICD-10-CM code. 

## getting CHARS counts by `detailed` ICD-CM category

```{r}
  mydetailed <- chars_icd_ccs_count(ph.data = charsDT, 
                                detailed = '^hypertension$')
```

```{r}
#| echo: FALSE
  mydetailed[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mydetailed, format = "markdown") 
```

## getting CHARS counts by `midlevel` ICD-CM category

```{r}
  mymidlevel <- chars_icd_ccs_count(ph.data = charsDT, 
                                midlevel = '^Hypertension$')
```

```{r}
#| echo: FALSE
  mymidlevel[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mymidlevel, format = "markdown") 
```

## getting CHARS counts by `broad` ICD-CM category

```{r}
  mybroad <- chars_icd_ccs_count(ph.data = charsDT, 
                                broad = 'Diseases of the circulatory system')
```

```{r}
#| echo: FALSE
  mybroad[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mybroad, format = "markdown") 
```

## getting CHARS counts by `superlevel` ICD-CM category

```{r}
  mysuperlevel <- chars_icd_ccs_count(ph.data = charsDT, 
                                superlevel = 'Chronic diseases')
```

```{r}
#| echo: FALSE
  mysuperlevel[, hospitalizations := format(hospitalizations, big.mark = ',')]
  kable(mysuperlevel, format = "markdown") 
```

# chars_injury_matrix()

The `chars_injury_matrix()` function provides a handy reference of all the `mechanism` and `intent` combinations that can be used with `chars_injury_matrix_count()`. Here are the first 10 rows:

```{r}
#| echo: FALSE
  kable(chars_injury_matrix()[1:10], format = "markdown") 
```

If you just want to see a list of the available intents, type `unique(chars_injury_matrix()[]$intent)`:

```{r}
#| echo: FALSE
sort(as.character(unique(chars_injury_matrix()[]$intent)))
```

Similarly, to see the available mechanisms, type `unique(chars_injury_matrix()[]$mechanism)`:

```{r}
#| echo: FALSE
sort(as.character(unique(chars_injury_matrix()[]$mechanism)))
```

# chars_injury_matrix_count()

The `chars_injury_matrix_count()` function is similar to the `chars_icd_ccs_count()` function above, except that it counts injury related hospitalizations. `chars_injury_matrix_count()` takes seven potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of CHARS data downloaded with `apde.data::chars()` or structured like `rads.data::synthetic_chars`. Note that the intents and mechanisms are pre-calculated so you will need to ensure `ph.data` has the relevant `injury_mechanism` and `injury_intent` columns. The easiest way to do this with real data is to have `apde.data::chars()` download all the columns.
-   `intent`: the injury intent of interest. Partial strings are allowed. Use `'none'` or `'any'` to ignore intent and return "Any intent". Use `'*'` (the default wildcard) to return all possible intents.
-   `mechanism`: the injury mechanism of interest. Partial strings are allowed. Use `'none'` or `'any'` to ignore mechanism and return "Any mechanism". Use `'*'` (the default wildcard) to return all possible mechanisms.
-   `group_by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `def`: acceptable values are 'narrow' or 'broad'. It specifies whether you want to use the CDC's recommended 'narrow' approach, which requires that the **principal diagnosis** of an injury hospitalization be a nature-of-injury ICD-10-CM code. Or, alternatively, the 'broad' definition that searches all available diagnosis fields on the hospital discharge record. See [this document](https://kc1.sharepoint.com/teams/DPH-APDEData/Shared%20Documents/Forms/AllItems.aspx?id=%2Fteams%2FDPH%2DAPDEData%2FShared%20Documents%2FCHARS%2FAPDE%20injury%20hospitalization%20method%5F2023%2Epdf&parent=%2Fteams%2FDPH%2DAPDEData%2FShared%20Documents%2FCHARS) for details.
-   `primary_ecode`: logical (T\|F) specifying whether to limit the analysis to using just the primary ecode (i.e., the `injury_ecode` variable), rather than all available ecodes. The vast majority of the time you will want to keep the default setting.
-   `kingco`: logical (T\|F) specifying whether to limit the data analysis to King County. Only works if ph.data still has the `chi_geo_kc` column.

## Specifying a single intent and ignoring the mechanism

```{r}
  mat1 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'assault', 
                              mechanism = 'none')
```

```{r}
#| echo: FALSE
    kable(mat1, format = "markdown")
```

## Specifying more than one intent and ignoring the mechanism

```{r}
#| warning: false
#| message: false
  mat2 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'assault|undetermined', 
                              mechanism = 'none')
```

```{r}
#| echo: FALSE
    kable(mat2, format = "markdown")
```

Note that you can also specify more than one intent or mechanism using a vector with separated values.

```{r}
#| warning: false
#| message: false
  mat2.alt <- chars_injury_matrix_count(ph.data = charsDT, 
                                        intent = c('assault', 'undetermined'), 
                                        mechanism = 'none')
```

## Specifying a single mechanism and ignoring the intent

```{r}
#| warning: false
#| message: false
  mat3 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'none', 
                              mechanism = 'motor_vehicle_traffic')
```

```{r}
#| echo: FALSE
    kable(mat3, format = "markdown")
```

## What happens if you specify 'none' or 'any' for both the mechanism and intent?

You get hospitalizations due to any injury.

```{r}
#| warning: false
#| message: false
  mat4 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'none', 
                              mechanism = 'none')
```

```{r}
#| echo: FALSE
    mat4[, hospitalizations := format(hospitalizations, big.mark = ',')]
    kable(mat4, format = "markdown")
```

## What happens if you don't specify the mechanism and intent?

You get every possible combination of mechanism and intent. Let's look at just the top 10 for convenience.

```{r}
#| warning: false
#| message: false
  mat5 <- chars_injury_matrix_count(ph.data = charsDT)[1:10]
```

```{r}
#| echo: FALSE
    mat5[, hospitalizations := format(hospitalizations, big.mark = ',')]
    kable(mat5[1:10], format = "markdown")
```

## How different are the `narrow` and `broad` definitions?

```{r}
#| warning: false
#| message: false
  mat6 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'none', 
                              mechanism = 'none', 
                              def = 'narrow')

  mat7 <- chars_injury_matrix_count(ph.data = charsDT, 
                              intent = 'none', 
                              mechanism = 'none', 
                              def = 'broad')
  
  deftable <- rbind(cbind(def = 'narrow', mat6),
                    cbind(def = 'broad', mat7))
```

```{r}
#| echo: FALSE
deftable[, hospitalizations := format(hospitalizations, big.mark = ',')]

    kable(deftable, format = "markdown")

```

These tables show that there is a huge difference in the number of hospitalizations, dependent upon the definition that you use. Unless you have a specific rationale for changing it, please use the default in your analyses (i.e., `def = 'narrow'`).

# Conclusion

We know this was a lot to process. The good news is that this vignette isn't going anywhere. If you remember (a) that this vignette exists and (b) where to find it, you'll be in good shape to take on standard CHARS analyses in the future.

If you've read through this vignette and the corresponding help files and are still confused, please feel free to reach out for assistance. You may have found a bug, who knows? Good luck!

-- *\`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
