---
title: "Death Functions"
format:
  html:
    embed-resources: true
  gfm:
    prefer-html: false
vignette: >
  %\VignetteIndexEntry{Death Functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{knitr}
editor: source
---

# Introduction

The [rads package](https://github.com/PHSKC-APDE/rads/) has a suite of tools designed to facilitate and accelerate the analysis of standardized death data. Combining the `rads` functions below with the clean death data should allow APDE analysts to conduct custom analyses with relative ease.

The core `rads` death functions are:

-   `death_icd10_clean()`: clean and standardize ICD-10 codes for use with `rads` death functions
-   `death_validate_data()`: validates that a dataset meets the requirements for use with `rads` death functions.
-   `death_113_count()`: generate death counts for the CDC's 113 Selected Causes of Death (PLUS COVID-19)
-   `death_130_count()`: generate death counts for the CDC's 130 Selected Causes of **infant** Death (PLUS COVID-19)
-   `death_injury_matrix_count()`: generate counts of external cause (a.k.a., injury related) deaths by intent and mechanism
-   `death_multicause_count()`: generate death counts needing **BOTH** underlying and contributing causes
-   `death_other_count()`: generate counts for select causes of death that are **not** **included** in `death_113_count()`, `death_130_count()`, & `death_injury_matrix_count()`, but may be of interest for public health analyses
-   `life_table_prep()`: prepare dataset for use with the `life_table()` function
-   `life_table()`: generate a standard life table, where the first row is the life expectancy at birth (a.k.a. 'expectation of life')

All of these functions have detailed help files that are accessible by typing `?function_name`, e.g., `?death_113_count`. Some examples for how to use these functions are given below.

***A few quick notes before we begin ...***

-   `death_113_count()`, `death_130_count()`, `death_injury_matrix_count()`, `death_multicause_count()` & `death_other_count()` only work with ICD-10 cause of death codes (i.e., those used since 1999).
-   If you want to create age-adjusted rates, we recommend you read the [age_standardize](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) and [calculating_rates_with_rads](https://github.com/PHSKC-APDE/rads/wiki/calculating_rates_with_rads) vignettes after reading through this one.
-   To perform a leading causes of death analysis, please follow the [Six Step Process on the wiki](https://github.com/PHSKC-APDE/rads/wiki/leading_causes_of_death)

# Set up the environment

```{r}
#| warning: false
#| message: false
rm(list=ls())
library(rads)
library(rads.data) # installed automatically when you installed rads
library(data.table)
```

```{r}
#| echo: false
pretty_kable <- function(dt) {
  knitr::kable(dt, format = 'markdown')
}
```

# Importing death data

For this wiki, we will use a prepared synthetic dataset (`rads.data::synthetic_death`). However, the tools are generalized and designed to be usable with any well formatted table of death data.

```{r}
#| warning: false
#| message: false
deaths <- rads.data::synthetic_death
names(deaths)
summary(deaths$age)
summary(deaths$date_of_death)
table(deaths$temperament)
```

Our synthetic death data has one column for the underlying cause of death (`underlying_cod_code`) and 20 for the contributing causes of death (`record_axis_code_`*). It is top coded at age 100, spans from 2019-2023, and has death and birth dates. If your data set has equivalent columns, you should be able to use all of the death functions. In addition, the synthetic data has a categorical 'temperament' variable that we will use to demonstrate stratification below. 

# death_icd10_clean()

`death_icd10_clean()` takes a single argument:

-   `icdcol`: the name of a character vector of ICD-10 codes.

The function can be used to clean any arbitrary vector of ICD-10 codes. Notice that it gives a warning, informing you that the codes were cleaned.

```{r}
#| echo: true
icd_codes <- c('G30.9', 'I25.1', 'I250', 'C349', 'U07-1', 'X44')
cleaned_icd_codes <- death_icd10_clean(icd_codes)
print(cleaned_icd_codes)
```

The function can also be used within a data.table. Notice that this time it gives a different warning, informing you that it had to replace one of the values (i.e., '3X44') with NA because it did not appear to be an ICD-10 code.

```{r}
#| echo: true
#| results: 'hide'
mydt <- data.table(icd_codes = c('G309', 'I251', 'I250', 
                                 'C349', 'U071', '3X44'))
mydt[, cleaned_icd_codes := death_icd10_clean(icd_codes)]

```

```{r}
#| echo: false
pretty_kable(head(mydt))
```

# death_validate_data()
`death_validate_data()` is used to ensure that your dataset is in good shape for use with the `rads::death_*_clean()` functions. It takes five arguments:

- `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
- `icdcol`: name of the column with the ICD codes of interest. Defaults to `"underlying_cod_code"`, which is used in APDE's data.
- `check_multicause`: Do you also want to validate the contributing cause of death columns? Defaults to `FALSE`, which means validation is limited to `icdcol`.	
- `contributing_cols`: The stem name of the contributing cause of death columns in `ph.data`. Only used when `check_multicause = TRUE`. The default is `contributing_cols = 'record_axis_code'`, which means it will validate `record_axis_code_1`, `record_axis_code_2`, etc.
- `verbose`: When `TRUE` (default), prints informational messages about validation results. When `FALSE`, only shows warnings and errors.

For many analyses, you will only need to provide the name of your dataset (`ph.data`) and the name of the column with the underlying cause of death ICD-10 code (`icdcol`). We want to demonstrate the use of the `death_multicause_count()` function below, so we will set `check_multicause = TRUE` and `contributing_cols = "record_axis_code"`.

If there are issues, it will generate informative errors or warnings. 

```{r}
death_validate_data(
  ph.data = deaths,
  icdcol = "underlying_cod_code",
  check_multicause = TRUE,
  contributing_cols = "record_axis_code",
  verbose = TRUE
)
```

If you didn't see an error or warning or notification, that means your data is ready to go.


# death_113_count()

`death_113_count()` allows the user to get death counts following the CDC's National Center for Health Statistics (NCHS) National Vital Statistics System's (NVSS) list of 113 selected causes of death ([see Table B](https://www.cdc.gov/nchs/data/dvs/Part9InstructionManual2020-508.pdf)). To review this list, simply type `death_113()` into your R console. You can use this information to select information for either the `causeids` or `cause` arguments of `death_113_count()`.

Here is a snapshot of the top 10 rows from `death_113()`

```{r}
#| echo: false
nchs113 <- death_113()[1:10]
pretty_kable(nchs113)
```

`death_113_count()` takes eight potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes.
-   `causeids`: specifies the causeid(s) for NCHS 113 Causes of Death (1:113).
-   `cause`: a character vector specifying the causes of death of interest. When specified, it is used in lieu of the `causeids`. It is case insensitive and matches partial strings.
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to `"underlying_cod_code"`, which is used in APDE's data.
-   `by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #1: Count 2020 deaths from COVID-19 or viral hepatitis

Count deaths due to COVID-19 or viral hepatitis

```{r}
#| warning: false
#| message: false
deaths.20 <- deaths[between(date_of_death, 
                            as.Date('2020-01-01'), 
                            as.Date('2020-12-31'))]

dt1 <- death_113_count(ph.data = deaths.20,
                       cause = 'viral hep|covid')
```

```{r}
#| echo: false
dt1[, deaths := format(deaths, big.mark = ',')]
pretty_kable(head(dt1))
```

Glancing at the table above, you'll see that you have the two causes of death of interest, PLUS the total of death from any cause ('All causes').

We could have specified 'Viral hepatitis' using `causeids` rather than the `cause` text string for the same result. Also, we really can't specify COVID-19. Since COVID-19 is not on the official 113 causes of death list, it does not have a code. It is however returned automatically by this function.

```{r}
#| warning: false
#| message: false
dt1.alt <- death_113_count(ph.data = deaths.20,
                           causeids = 14)
```

```{r}
#| echo: false
dt1.alt[, deaths := format(deaths, big.mark = ',')]
pretty_kable(head(dt1.alt))
```

### example #2: 2019 & 2020 top five causes of death in `death_113_count()`

First, let's subset to the 2019 & 2020 data ...

```{r}
#| warning: false
#| message: false
deaths.19_20 <- deaths[between(date_of_death, 
                               as.Date('2019-01-01'), 
                               as.Date('2020-12-31'))]

deaths.19_20[, year := year(date_of_death)]
```

Now, let's use `by` to get the 113 cause of death counts by year

```{r}
#| warning: false
#| message: false
dt113_19_20 <- death_113_count(ph.data = deaths.19_20,
                               by = 'year')
```

Finally, let's select the top five causes of death by year using data.table

```{r}
#| warning: false
#| message: false
dt113_19_20[cause.of.death != 'All causes', 
            rank := frank(-deaths), 
            by = 'year']
dt113_19_20 <- dt113_19_20[rank <= 5 ]
setorder(dt113_19_20, year, rank)
```

```{r}
#| echo: false
dt113_19_20[, deaths := format(deaths, big.mark = ',')]
pretty_kable(dt113_19_20)
```

***Note!*** Calculating the leading causes of death that we report publicly is more complicated. Please refer to the [wiki](https://github.com/PHSKC-APDE/rads/wiki/leading_causes_of_death) and [CHI repository README](https://github.com/PHSKC-APDE/chi/tree/main/death#leading-causes-of-death) for details.

### example #3: Calculating years of potential life lost due to COVID-19 in 2020, assuming age of death at 65 as the standard

First, identify deaths due to COVID-19. COVID-19 is not one of the official 113 causes of death. So let's arbitrarily get causeid #1 (Salmonella infections), knowing that COVID-19 is always returned in addition to your specified cause(s).

```{r}
#| warning: false
#| message: false
dt3 <- death_113_count(ph.data = deaths.20,
                       causeids = 1,
                       death_age_col = 'age',
                       ypll_age = 65)
dt3 <- dt3[cause.of.death == 'COVID-19 (U07.1)'] # limit output to COVID-19
```

```{r}
#| echo: false
dt3[, deaths := format(deaths, big.mark = ',')]
dt3[, ypll_65 := format(ypll_65, big.mark = ',')]
pretty_kable(head(dt3))
```

The table above tells us that, if everyone who died from COVID-19 in 2020 had lived to at least age 65, they would have lived a total of `r dt3$ypll_65` additional years.

# death_130_count()

`death_130()` and `death_130_count()` are used for the [NCHS 130 Selected Causes of Infant Death (Table C)](https://www.cdc.gov/nchs/data/dvs/Part9InstructionManual2020-508.pdf). They work in exactly the same way as `death_113()` and `death_113_count()` and will therefore not be described in detail.

# death_injury_matrix_count()

`death_injury_matrix_count()` allows the user to get injury related death counts by intent and mechanism. To review a table of all available combinations of mechanism and intent, simply type `death_injury_matrix()` into your R console. Here are the first six lines ...

```{r}
  mech.n.intent <- death_injury_matrix()[1:6]
```

```{r}
#| echo: false
  pretty_kable(head(mech.n.intent))
```

`death_injury_matrix_count()` takes eight potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
-   `intent`: specifies the intent(s) of interest as per options available in `death_injury_matrix()`. `'none'` will ignore intent and return 'Any intent'. `'\*'` is a wildcard for all possible intents.
-   `mechanism`: specifies the mechanism(s) of interest as per options available in `death_injury_matrix`. `'none'` will ignore mechanism and return 'Any mechanism'. `'\*'` is a wildcard for all possible mechanisms.
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to `"underlying_cod_code"`, which is used in APDE's data.
-   `by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #4: 2019-2020 injury related deaths (every combination of intent and mechanism), stratified by temperament

Run `death_injury_matrix_count()` and show the six random rows

```{r}
#| message: false
#| warning: false
dt4 <- death_injury_matrix_count(ph.data = deaths.19_20,
                                 intent = "*",
                                 mechanism = "*",
                                 by = 'temperament')
```

```{r}
#| echo: false
dt4[, deaths := format(deaths, big.mark = ',')]
pretty_kable(dt4[deaths >10])
```

**What happens if you set both intent & mechanism to "none"?** The function assumes you want the total injury related deaths (without regard to intent or mechanism).

```{r}
#| message: false
#| warning: false
dt4.alt <- death_injury_matrix_count(ph.data = deaths.19_20,
                                     intent = "none",
                                     mechanism = "none",
                                     by = 'temperament')
```

```{r}
#| echo: false
dt4.alt[, deaths := format(deaths, big.mark = ',')]
pretty_kable(head(dt4.alt))
```

### example #5: 2019-2023 homicides (intent) by firearms (mechanism), stratified by temperament

Run `death_injury_matrix_count`

```{r}
#| message: false
#| warning: false
deaths.19_23 <- deaths[between(date_of_death, 
                               as.Date('2019-01-01'), 
                               as.Date('2023-12-31'))]

deaths.19_23[, year := year(date_of_death)]

dt5 <- death_injury_matrix_count(ph.data = deaths.19_23,
                                 intent = "homicide",
                                 mechanism = "firearm",
                                 by = 'temperament')
```

```{r}
#| echo: false
pretty_kable(dt5)
```

### example #6: years of potential life lost due to firearm homicides 2019-2023

Run `death_injury_matrix_count`

```{r}
#| message: false
#| warning: false
dt6 <- death_injury_matrix_count(ph.data = deaths.19_23,
                                 intent = "homicide",
                                 mechanism = "firearm",
                                 death_age_col = 'age',
                                 ypll_age = 65)
```

```{r}
#| echo: false
dt6[, ypll_65 := format(ypll_65, big.mark = ',')]
pretty_kable(dt6)
```

# death_multicause_count()

`death_multicause_count()` allows the user to get counts of causes of death requiring **BOTH** *underlying* **AND** *contributing* causes. To view available analytic ready multicause death categories, simply type `death_multicause()`. If you want to add more causes of death to this list, [please submit a GitHub issue](https://github.com/PHSKC-APDE/rads/issues/new?template=issue_form.yml). In the meantime the function will allow you to specify the underlying cause of death and contributing cause of death ICD-10 of interest.

```{r}
#| echo: true
#| results: 'hide'
death_multicause()
```

```{r}
#| echo: false
pretty_kable(death_multicause())
```

`death_multicause_count()` takes eleven potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes, with at least one column of ICD10 underlying cause codes and columns for contributing cause codes.
-   `cause_name`: a character vector specifying the multicause death definition from the reference table. For example, `cause_name = "Opioid"`. To see available options, use `death_multicause()`.
-   `underlying_codes`: a character vector of ICD10 codes that must appear in the underlying cause of death. Only used if `cause_name = NULL`. For example, `c("X40", "X41", "X42")`.
-   `contributing_codes`: a character vector of ICD10 codes that must appear in the contributing causes of death. Only used if `cause_name = NULL`. For example, `c("T400", "T401")`.
-   `icdcol`: name of the column with the underlying cause of death ICD codes. Defaults to `"underlying_cod_code"`, which is used in APDE's data.
-   `contributing_cols`: a character vector specifying the stem of the column names containing contributing cause codes. The function expects columns named with this stem followed by numbers 1–20. The default, `"record_axis_code"`, leads the function to expect columns named `'record_axis_code_1'`, `'record_axis_code_2'`, ... `'record_axis_code_20'`.
-   `contributing_logic`: a character vector specifying whether `"ANY"` (default) or `"ALL"` of the `contributing_codes` must be present.
-   `by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #7: Count the number of opioid deaths for 2019-2023, stratified by temperament

```{r}
#| message: false
#| warning: false
dt7 <- death_multicause_count(ph.data = deaths.19_23,
                                 cause_name = "Opioid",
                                 by = "temperament")
dt7 <- dt7[cause.of.death != 'All causes']
```

```{r}
#| echo: false
dt7[, deaths := format(deaths, big.mark = ',')]
pretty_kable(dt7)
```

# death_other_count()

`death_other_count()` allows the user to get death counts for cause of death that are NOT included in `death_113_count()`, `death_130_count()`, and `death_injury_matrix_count()`, but that are still of general interest in public health analyses. To review a list of all available 'other' causes of death, simply type `death_other()` into your R console. If you want to add additional causes of death to this list, [please submit a GitHub issue](https://github.com/PHSKC-APDE/rads/issues/new?template=issue_form.yml).

```{r}
death_other()
```

`death_other_count()` takes seven potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
-   `cause`: a character vector specifying the cause(s) of death of interest taken from the list provided by `death_other()`. It is case insensitive and matches partial strings.
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to `"underlying_cod_code"`, which is used in APDE's data.
-   `by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #8: Count the number of influenza / pneumonia deaths for 2019-2023, stratified by year

```{r}
#| message: false
#| warning: false
dt8 <- death_other_count(ph.data = deaths.19_23,
                         cause = "pneumonia",
                         by = "year")
dt8 <- dt8[cause.of.death != 'All causes']
setorder(dt8, year, cause.of.death)
```

```{r}
#| echo: false
dt8[, deaths := format(deaths, big.mark = ',')]
pretty_kable(dt8)
```

### example #9: Count the number of cancer and heart disease deaths in 2021

Run `death_other_count`

```{r}
#| message: false
#| warning: false
deaths.21 <- deaths[between(date_of_death, 
                            as.Date('2021-01-01'), 
                            as.Date('2021-12-31'))]

dt9 <- death_other_count(ph.data = deaths.21,
                         cause = "cancer|heart",
                         death_age_col = 'age',
                         ypll_age = 65)
```

```{r}
#| echo: false
dt9[, deaths := format(deaths, big.mark = ',')]
dt9[, ypll_65 := format(ypll_65, big.mark = ',')]
pretty_kable(dt9)
```

# life_table()

`life_table()` generates a standard life table, where the first row is the life expectancy (a.k.a. 'expectation of life') at birth. Since it needs aggregate death data, you are **strongly encouraged to preprocess the death data with `life_table_prep()`.** A simple example of how to use `life_table_prep()` is embedded below.

`life_table()` takes seven potential arguments:

-   `ph.data`: the name of data.table/data.frame with aggregated deaths and corresponding populations, as well as the age interval and the average fraction of years lived in the interval by those who die in the interval.
-   `myages`: the name of a column in `ph.data` with the age intervals used for the life table calculations (e.g., c('0-1', '1-5', ... '85+')).
-   `mydeaths`: the name of a numeric column in `ph.data` with the total deaths for the given age interval in the given year.
-   `mypops`: the name of a numeric column in `ph.data` with the total population corresponding to `mydeaths`.
-   `myprops`: the name of a numeric column in `ph.data` with the average proportion of the interval lived by those who died in the interval.
-   `by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ci`: the confidence interval, must be \[0, 1\]. Default is 0.95

Please refer to the help file for more details.

### example #10: Get life expectancy at birth for residents born 2019-2023, stratified by year

Use `life_table_prep()` to aggregate the age interval specific count of deaths and the average fraction of the interval lived by those who died in that interval, stratified by year of death. **Note**, by default `life_table_prep()` uses the standard age intervals used by the CDC and WA DOH. These intervals can easily be customized if desired. Here we show just the top six rows so you can see the data structure.

```{r}
#| message: false
#| warning: false
dt10 = life_table_prep(ph.data = deaths.19_23, 
                       dobvar = "date_of_birth",
                       dodvar = "date_of_death",
                       by = 'year')
```

```{r}
#| echo: false
pretty_kable(head(dt10))
```

Filter/delete rows with missing ages because they would not have a corresponding population.

```{r}
#| message: false
#| warning: false
dt10 <- dt10[!is.na(ages)]
```

The death data is now ready for `life_table`, but we still need to merge on the corresponding populations. Let's start by getting the raw population for the same time period (2019–2023) by year and age.

```{r}
#| message: false
#| warning: false
pop <- rads.data::synthetic_population
pop <- pop[year %in% 2019:2023]
pop <- pop[, .(pop = sum(pop)), .(year, age)]
```

```{r}
#| echo: false
setorder(pop, age)
pretty_kable(head(pop))
```

Now we need to aggregate the population data to use the same age intervals as the death data.

```{r}
#| message: false
#| warning: false
my.cuts <- c(0, 1, 5, 10, 15, 18, 20, 25, 30, 35, 40, 
             45, 50, 55, 60, 65, 70, 75, 80, 85)
pop[, ages := cut(age, my.cuts, right = F)] # create age intervals
pop[, ages := gsub("\\[|\\]|\\)", "", ages)] # tidy
pop[, ages := gsub("\\,", "-", ages)] #tidy
pop[age >= max(my.cuts), ages := paste0(max(my.cuts), "+")] # tidy
pop <- pop[, .(pop = sum(pop)), .(year, ages)] #tidy
pop[, year := as.integer(year)]
```

```{r}
#| echo: false
pretty_kable(head(pop))
```

Merge the properly formatted death and population data together.

```{r}
#| message: false
#| warning: false
dt10 <- merge(dt10, 
              pop, 
              by = c('year', 'ages'), 
              all = T)
```

```{r}
#| echo: false
pretty_kable(head(dt10))
```

Run `life_table()` function to create life tables by year.

```{r}
#| message: false
#| warning: false
dt10.lifetable <- life_table(ph.data = dt10, 
                             by = 'year')
```

Display the first row of the life table for each year, which is for children aged 0-1. The `ex` column is the expectation of life, which for 0-1 year olds is life expectancy at birth. To see a detailed key of what all the columns mean, type `?life_table` in your R console and read the 'Details' section.

```{r}
#| message: false
#| warning: false
#| results: 'hide'
dt10.lifetable <- dt10.lifetable[ages == '0-1',]
dt10.lifetable <- dt10.lifetable[, .(year, ages, ex, ex_lower, ex_upper, ex_se)]
```

```{r}
#| echo: false
pretty_kable(dt10.lifetable)
```

***Note!** The life expectancy values in the table above are irrationally high because we used synthetic data, which do not accurately reflect the age at death in the original population.* 

# Conclusion

We know this was a lot to process. The good news is that this vignette isn't going anywhere. If you remember (a) that this vignette exists and (b) where to find it, you'll be in good shape to take on standard mortality analyses in the future.

If you've read through this vignette and the corresponding help files and are still confused, please feel free to reach out for assistance. You may have found a bug, who knows? Good luck!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
