---
title: "Death Functions"
format: gfm
prefer-html: false
self-contained: true
editor: visual
---

# Introduction

The [rads package](https://github.com/PHSKC-APDE/rads/) has a suite of tools designed to facilitate and accelerate the analysis of standardized death data. Combining the `rads` functions below with the clean death data on our servers should allow APDE analysts to conduct custom analyses with relative ease.

The core `rads` death function are:

-   `get_data_death()`: easily download standardized death data from SQL into R
-   `death_113_count()`: generate death counts for the CDC's 113 Selected Causes of Death (PLUS COVID-19)
-   `death_130_count()`: generate death counts for the CDC's 130 Selected Causes of **infant** Death (PLUS COVID-19)
-   `death_injury_matrix_count()`: generate counts of external cause (a.k.a., injury related) deaths by intent and mechanism
-   `death_other_count()`: generate counts for select causes of death that are **not** **included** in `death_113_count()`, `death_130_count()`, & `death_injury_matrix_count()`, but are of interest to APDE.
-   `life_table()`: generate a standard life table, where the first row is the life expectancy (aka 'expectation of life') at birth

All of these functions have detailed help files that are accessible by typing `?function_name`, e.g. `?get_data_death`. Some examples for how to use these functions are given below.

***A few quick notes before we begin ...***

-   `death_113_count()`, `death_130_count()`, `death_injury_matrix_count()`, & `death_other_count()` only work with ICD-10 cause of death codes (i.e., those used since 1999).
-   If you want to create age-adjusted rates, we recommend you read the [age_standardize](https://github.com/PHSKC-APDE/rads/wiki/age_standardize) and [calculating_rates_with_rads](https://github.com/PHSKC-APDE/rads/wiki/calculating_rates_with_rads) vignettes after reading through this one.

# Set up the environment

```{r}
#| warning: false
#| message: false
rm(list=ls())
library(rads)
library(data.table)
```

```{r}
#| echo: FALSE
library(kableExtra)
```

# get_data_death()

`get_data_death()` takes four potential arguments:

-   `cols`: the names of the columns that you want to download. You will dramatically speed up the download time if you only specify the columns of genuine interest.
-   `year`: the year(s) of interest, from 1980 to the present.
-   `kingco`: logical (T\|F) specifying whether to limit the download to King County.
-   `topcode`: logical (T\|F) specifying whether you want to top code `chi_age` at 100 to match population data.

If you do not specify any of the arguments, you will get all death data columns, for King County, for the latest year, with top coding to 100.

```{r}
#| warning: false
#| message: false
ex0 <- get_data_death()
ncol(ex0) # number of columns in the downloaded death data  
names(ex0)[1:6] # names of the first 6 columns
unique(ex0$chi_geo_kc) # confirm data is limited to King County
unique(ex0$chi_year) # check the year
max(ex0$chi_age, na.rm = T) # check top coding
```

# death_113_count()

`death_113_count()` allows the user to get death counts following the CDC's National Center for Health Statistics (NCHS) National Vital Statistics System's (NVSS) list of 113 selected causes of death ([see Table B](https://www.cdc.gov/nchs/data/dvs/Part9InstructionManual2020-508.pdf)). To review this list, simply type `death_113()` into your R console. You can use this information to select information for either the `causeids` or `cause` arguments of `death_113_count()`.

Here is a snapshot of the top 10 rows from `death_113()`

```{r}
#| echo: FALSE
  nchs113 <- death_113()[1:10]
  kable((nchs113), format = "markdown") 
```

`death_113_count()` takes eight potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
-   `causeids`: specifies the causeid(s) for NCHS 113 Causes of Death (1:113)
-   `cause`: optional. A character vector specifying the causes of death of interest. When specified, it is used in lieu of the `causeids`. It is case insensitive and matches partial strings.
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to "underlying_cod_code", which is used in APDE's data.
-   `kingco`: logical (T\|F) specifying whether to limit the data analysis to King County. Only works if ph.data still has the `chi_geo_kc` column.
-   `group_by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: optional. The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: optional. Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #1: Count 2020 deaths from COVID-19 or viral hepatitis

Get the 2020 data from SQL

```{r}
ex1 <- get_data_death(year = 2020, 
                      cols = c('chi_year', 'chi_geo_kc',
                               'underlying_cod_code'))
```

Check the dimensions and column names

```{r}
dim(ex1)
names(ex1)
```

Lets take a peak at the first 10 rows where the underlying cause of death is not missing to see the data structure.

```{r}
#| echo: FALSE
kable((ex1[!is.na(underlying_cod_code)][1:10]), format = "markdown") 
```

Count deaths due to COVID-19 or viral hepatitis

```{r}
#| warning: false
#| message: false
dt1 <- death_113_count(ph.data = ex1, 
                       cause = 'viral hep|covid')
```

```{r}
#| echo: false
dt1[, deaths := format(deaths, big.mark = ',')]
kable(head(dt1), format = "markdown")
```

Glancing at the table above, you'll see that you have the two causes of death of interest, PLUS the total of death from any cause ('All causes') PLUS any rows where the ICD code did not match the selected causes of death ('Missing/Unknown'). *Missing/Unknown = All causes - Viral hepatitis - COVID-19.*

We could have specified 'Viral hepatitis' using `causeids` rather than the `cause` text string for the same result. Also, we really can't specify COVID-19. Since COVID-19 is not on the official 113 causes of death list, it does not have a code. It is however returned automatically by this function.

```{r}
#| warning: false
#| message: false
dt1.alt <- death_113_count(ph.data = ex1, 
                       causeids = 14)
```

```{r}
#| echo: false
dt1.alt[, deaths := format(deaths, big.mark = ',')]
kable(head(dt1.alt), format = "markdown")
```

### example #2: 2019 & 2020 top five causes of death

First, let's get 2019 & 2020 data from SQL ...

```{r}
#| warning: false
#| message: false
ex2 <- get_data_death(year = 2019:2020, 
                      cols = c('chi_year', 'chi_geo_kc', 
                               'underlying_cod_code'))
```

Now, let's use `group_by` to get counts by year

```{r}
#| warning: false
#| message: false
dt113_19_20 <- death_113_count(ph.data = ex2, 
                               group_by = 'chi_year')
```

Finally, let's select the top five causes of death by year using data.table

```{r}
#| warning: false
#| message: false
dt113_19_20[cause.of.death != 'All causes', 
            rank := frank(-deaths), 
            by = 'chi_year']
dt113_19_20 <- dt113_19_20[rank <=5 ]
setorder(dt113_19_20, chi_year, rank)
```

```{r}
#| echo: false
dt113_19_20[, deaths := format(deaths, big.mark = ',')]
kable(dt113_19_20, format = "markdown")
```

### example #3: Calculating years of potential life lost due to COVID-19, assuming age of death at 65 as the standard

Again, let's load 2020 death data.

```{r}
ex3 <- get_data_death(year = 2020, 
                      cols = c('chi_age', 'underlying_cod_code'))
```

Now, identify deaths due to COVID-19. Remember COVID-19 is not a one of the official 113 causes of death, so let's arbitrarily get causeid #1 (Salmonella infections), knowing that it will also give us COVID-19.

```{r}
#| warning: false
#| message: false
dt3 <- death_113_count(ph.data = ex3, 
                       causeids = 1,
                       kingco = FALSE,
                       ypll_age = 65)
dt3 <- dt3[cause.of.death == 'COVID-19 (U07.1)'] # limit output to COVID-19
```

```{r}
#| echo: false
dt3[, deaths := format(deaths, big.mark = ',')]
dt3[, ypll_65 := format(ypll_65, big.mark = ',')]
kable(head(dt3), format = "markdown")
```

The table above tells us that, if everyone who died from COVID-19 in 2020 had lived to at least age 65, they would have lived a total of `r dt3$ypll_65` additional years.

# death_130_count()

`death_130()` and `death_130_count()` work in exactly the same way as `death_113()` and `death_113_count()` and will therefore not be described in detail.

# death_injury_matrix_count()

`death_injury_matrix_count()` allows the user to get injury related death counts by intent and mechanism. To review a table of all available combinations of mechanism and intent, simply type `death_injury_matrix()` into your R console. Here are the first six lines ...

```{r}
  mech.n.intent <- death_injury_matrix()[1:6]
```

```{r}
#| echo: false
  kable(head(mech.n.intent), format = "markdown")
```

`death_injury_matrix_count()` takes eight potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
-   `intent`: specifies the intent(s) of interest as per options available in `death_injury_matrix`. 'none' will ignore intent and return 'Any intent'. '\*' is a wildcard for all possible intents.
-   `mechanism`: specifies the mechanism(s) of interest as per options available in `death_injury_matrix`. 'none' will ignore mechanism and return 'Any mechansim'. '\*' is a wildcard for all possible mechanisms.
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to "underlying_cod_code", which is used in APDE's data.
-   `kingco`: logical (T\|F) specifying whether to limit the data analysis to King County. Only works if ph.data still has the `chi_geo_kc` column.
-   `group_by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: optional. The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: optional. Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #4: 2020 King County injury related deaths (every combination of intent and mechanism), stratified by sex

Get the data

```{r}
ex4 <- get_data_death(year = 2020, 
                      cols = c('chi_sex', 'underlying_cod_code'))
```

Run death_injury_matrix_count and show the first six rows

```{r}
#| message: false
#| warning: false
dt4 <- death_injury_matrix_count(ph.data = ex4, 
                                 intent = "*", 
                                 mechanism = "*",
                                 kingco = FALSE, 
                                 group_by = 'chi_sex')[1:6]
```

```{r}
#| echo: false
dt4[, deaths := format(deaths, big.mark = ',')]
kable(dt4, format = "markdown")
```

Notice that you did not have to specify that you wanted the results stratified by sex. The function will automatically stratify by whatever extra columns are contained in `ph.data`. Be mindful of this and be sure that you submit `ph.data` with the minimum necessary number of columns.

**Now, what happens if you set both intent & mechanism to "none"?** The function assumes you want the total injury related deaths (without regard to intent or mechanism).

```{r}
#| message: false
#| warning: false
dt4.alt <- death_injury_matrix_count(ph.data = ex4, 
                                 intent = "none", 
                                 mechanism = "none",
                                 kingco = FALSE, 
                                 group_by = 'chi_sex')
```

```{r}
#| echo: false
dt4.alt[, deaths := format(deaths, big.mark = ',')]
kable(head(dt4.alt), format = "markdown")
```

### example #5: 2016-2020 King County homicides (intent) by firearms (mechanism), stratified by sex

Get the data

```{r}
ex5 <- get_data_death(year = 2016:2020, 
                      cols = c('chi_sex', 'underlying_cod_code'))
```

Run death_injury_matrix_count

```{r}
#| message: false
#| warning: false
dt5 <- death_injury_matrix_count(ph.data = ex5, 
                                 intent = "homicide", 
                                 mechanism = "firearm", 
                                 kingco = FALSE, 
                                 group_by = 'chi_sex')
```

```{r}
#| echo: false
kable(dt5, format = "markdown")
```

### example #6: years of potential life lost due to firearm homicides 2016-2020

Get the data

```{r}
ex6 <- get_data_death(year = 2016:2020, 
                      cols = c('chi_age', 'underlying_cod_code'))
```

Run death_injury_matrix_count

```{r}
#| message: false
#| warning: false
dt6 <- death_injury_matrix_count(ph.data = ex6, 
                                 intent = "homicide", 
                                 mechanism = "firearm", 
                                 kingco = FALSE, 
                                 ypll_age = 65)
```

```{r}
#| echo: false
dt6[, ypll_65 := format(ypll_65, big.mark = ',')]
kable(dt6, format = "markdown")
```

# death_other_count()

`death_other_count()` allows the user to get death counts for cause of death that are NOT included in `death_113_count()`, `death_130_count()`, and `death_injury_matrix_count()`, but that are still of general interest to APDE. To review a list of all available 'other' causes of death, simply type `death_other()` into your R console. If you want to add additional causes of death to this list, [please submit a GitHub issue](https://github.com/PHSKC-APDE/rads/issues/new).

```{r}
death_other()
```

`death_other_count()` takes seven potential arguments:

-   `ph.data`: the name of a person level data.table/data.frame of death data with ICD codes
-   `cause`: a character vector specifying the cause(s) of death of interest taken from the list provided by `death_other()`. It is case insensitive and matches partial strings
-   `icdcol`: name of the column with the ICD codes of interest. Defaults to "underlying_cod_code", which is used in APDE's data
-   `kingco`: logical (T\|F) specifying whether to limit the data analysis to King County. Only works if ph.data still has the `chi_geo_kc` column.
-   `group_by`: identifies the variables by which you want to group (a.k.a., stratify) the results.
-   `ypll_age`: optional. The age in years (an integer) used for calculating Years of Potential Life Lost.
-   `death_age_col`: optional. Name of the age column used if `ypll_age` is specified.

Please refer to the help file for more details.

### example #7: Count the number of heat stress deaths for 2017-2021, stratified by year

Get the data

```{r}
ex7 <- get_data_death(year = 2017:2021, 
                      cols = c('chi_year', 'underlying_cod_code'))
```

Run `death_other_count`

```{r}
#| message: false
#| warning: false
dt7 <- death_other_count(ph.data = ex7, 
                                 cause = "heat", 
                                 kingco = FALSE, 
                                 group_by = "chi_year")
dt7 <- dt7[cause.of.death != 'All causes']
setorder(dt7, chi_year, cause.of.death)
```

```{r}
#| echo: false
dt7[, deaths := format(deaths, big.mark = ',')]
kable(dt7, format = "markdown")
```

### example #8: Count the number of cancer and heart disease deaths in 2021

Get the data

```{r}
ex8 <- get_data_death(year = 2021, 
                      cols = c('chi_age', 'underlying_cod_code'))
```

Run `death_other_count`

```{r}
#| message: false
#| warning: false
dt8 <- death_other_count(ph.data = ex8, 
                                 cause = "cancer|heart", 
                                 kingco = FALSE, 
                                 ypll_age = 65)
```

```{r}
#| echo: false
dt8[, deaths := format(deaths, big.mark = ',')]
dt8[, ypll_65 := format(ypll_65, big.mark = ',')]
kable(dt8, format = "markdown")
```

# life_table()

`life_table()` generates a standard life table, where the first row is the life expectancy (aka 'expectation of life') at birth. Since it needs aggregate death data, you are **strongly encouraged to preprocess the death data with `life_table_prep()`.**

`life_table()` takes six potential arguments:

-   `ph.data`: the name of data.table/data.frame with aggregated deaths and corresponding populations, as well as the age interval and the average fraction of years lived in the interval by those who die in the interval.
-   `myages`: the name of a column in `ph.data` with the age intervals used for the life table calculations (e.g., c('0-1', '1-5', ... '85+')).
-   `mydeaths`: the name of a numeric column in `ph.data` with the total deaths for the given age interval in the given year.
-   `mypops`: the name of a numeric column in `ph.data` with the total population corresponding to `mydeaths`.
-   `myprops`: the name of a numeric column in `ph.data` with the average proportion of the interval lived by those who died in the interval.
-   `ci`: the confidence interval, must be \[0, 1\]. Default is 0.95

Please refer to the help file for more details.

### example #9: Get life expectancy at birth for King County residents born 2016-2020, stratified by sex

Get the critical death data

```{r}
#| message: false
#| warning: false
ex9 <- get_data_death(year = 2016:2020, 
                        cols = c('date_of_birth', 'date_of_death', 'chi_sex'))
```

To preserve privacy, lets look at a summary of the table rather than the actual table

```{r}
#| message: false
#| warning: false
summary(ex9)
```

Use `life_table_prep()` to aggregate the age interval specific count of deaths and the average fraction of the interval lived by those who died in that interval, stratified by sex. **Note**, by default `life_table_prep()` uses the standard age intervals used by the CDC and WA DOH. These intervals can easily be customized if desired. Here we show just the top six rows so you can see the data structure.

```{r}
#| message: false
#| warning: false
dt9 = life_table_prep(DTx = ex9)
```

```{r}
#| echo: false
kable(head(dt9), format = "markdown")
```

Filter/delete rows with missing ages or non-binary sex because neither would have a corresponding population.

```{r}
#| message: false
#| warning: false
dt9 <- dt9[!is.na(chi_sex) & !is.na(ages)]
```

Rename `chi_sex` to `gender` so that it plays nicely with the population data below.

```{r}
#| message: false
#| warning: false
setnames(dt9, "chi_sex", "gender")
```

```{r}
#| echo: false
kable(head(dt9), format = "markdown")
```

The death data is now ready for `life_table`, but we still need to merge on the corresponding populations. Let's start by getting the raw population for the same time period (2016-2020) by gender and age.

```{r}
#| message: false
#| warning: false
pop <- get_population(kingco = T, 
                      years = 2016:2020, 
                      group_by = c('genders', 'ages'))
pop <- pop[, .(gender, age, pop)]
```

```{r}
#| echo: false
setorder(pop, age)
kable(head(pop), format = "markdown")
```

Now we need to aggregate the population data to use the same age intervals as the death data.

```{r}
#| message: false
#| warning: false
my.cuts <- c(0, 1, 5, 10, 15, 18, 20, 25, 30, 35, 40, 45, 50, 55, 60, 65, 70, 75, 80, 85)
pop[, ages := cut(age, my.cuts, right = F)] # create age intervals
pop[, ages := gsub("\\[|\\]|\\)", "", ages)] # tidy
pop[, ages := gsub("\\,", "-", ages)] #tidy
pop[age >= max(my.cuts), ages := paste0(max(my.cuts), "+")] # tidy
pop <- pop[, .(pop = sum(pop)), .(gender, ages)] #tidy
```

```{r}
#| echo: false
kable(head(pop), format = "markdown")
```

Merge the properly formatted death and population data together.

```{r}
#| message: false
#| warning: false
dt9 <- merge(dt9, 
             pop, 
             by = c('gender', 'ages'), 
             all = T)
```

```{r}
#| echo: false
kable(head(dt9), format = "markdown")
```

Run `life_table()` function to create life tables for males and females.

```{r}
#| message: false
#| warning: false
dt9.male <- life_table(ph.data = dt9[gender == 'Male'])
dt9.female <- life_table(ph.data = dt9[gender == 'Female'])
```

Display the first row of the life table, which is for children under 1. The `ex` column is the life expectancy at birth. To see a detailed key of what all the columns mean, type `?life_table` in your R console and read the 'Details' section.

```{r}
#| echo: false
kable(dt9.male[1,], format = "markdown")
kable(dt9.female[1,], format = "markdown")
```

# Conclusion

We know this was a lot to process. The good news is that this vignette isn't going anywhere. If you remember (a) that this vignette exists and (b) where to find it, you'll be in good shape to take on standard mortality analyses in the future.

If you've read through this vignette and the corresponding help files and are still confused, please feel free to reach out for assistance. You may have found a bug, who knows? Good luck!
