---
title: "get_population()"
output:
  github_document: default
  rmarkdown::html_vignette: default
  pdf_document: default
  urlcolor: blue
vignette: |
  %\VignetteEngine{knitr::knitr} %\VignetteIndexEntry{get_population}
---
```{r, echo = F, message=FALSE}
library(rads)
```
## Introduction

This vignette will provide some examples of ways to pull population data into R from the Azure cloud (thank you Jeremy!). Population numbers are estimated by the [WA Office of Financial Management (OFM) population unit](https://ofm.wa.gov/washington-data-research/population-demographics). OFM produces two sets of estimates: (1) [April 1 official population estimates](https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/april-1-official-population-estimates) for cities and towns and (2) [Small Area Estimates (SAE)](https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/small-area-estimates-program) for smaller geographies. The `get_population()` function pulls the SAE numbers and should be the same as those in [CHAT](https://secureaccess.wa.gov/doh/chat/Entry.mvc). 

**NOTE!!** To get the most out of this vignette, I highly recommend that you actually type each and every bit of code into R. Doing so will almost definitely help you learn the syntax much faster than just reading the vignette or copying and pasting the code.

## `get_population` arguments
Arguments are the values that we send to a function when it is called. Generally, typing `args(my_function_of_interest)` will return the possible arguments including any defaults. For example, 
```{r}
args(get_population)
```
The standard arguments for get_population() are:
 
1) `kingco` << a logical vector. Do you want the estimates limited to King County? The default is TRUE.

2) `years` << a numeric vector of the year(s) of interest. Currently only provides estimates for 2010+. The default is 2020.

3) `ages` << a numeric vector of the age(s) of interest. The acceptable range and current default is 0 to 100. (Note! Age 100 is actually the top coding for ages 100-120.)

4) `genders` << a character vector of the gender(s) of interest. The acceptable values are “f”, “female”, “m”, and “male”. The default is both female and male. 

5) `races` << a character vector of the racial/ethnic group(s) of interest. The acceptable values are "aian", "asian", "black", "hispanic", "multiple", "nhpi", and "white". The default is all the possible values. 

6) `race_type` << a character vector limited to “race” (Hispanic as an ethnicity) or “race_eth” (Hispanic as a race). The default is “race_eth”.

7) `geo_type` << a character vector describing the geographic level for which you want population estimates. Possible values are “kc”, "seattle, “blk”, “blkgrp”, “hra”, “region”, “tract”, and “zip”. Note that all these geo_types except "zip" are available for King, Pierce, and Snohomish counties only. The default geo_type is “kc”.

8) `group_by` << a character vector describing the how you would like to have the estimates grouped (i.e., stratified). For example, if you set the years argument to (2017:2019) and set the `group_by` argument to “years”, you would receive estimates for 2017, 2018, and 2019. Otherwise you would receive one estimate for 2017 through 2019. Valid options are limited to: "years", "ages", "genders", "race", "race_eth", "fips_co", and "geo_id". The default is NULL, meaning estimates are not grouped / stratified.

9) `round` << a logical vector. Do you want to round your population estimates to whole numbers? Default is TRUE. 

10) `mykey` << a character vector with the name of the `keyring::` key that provides access to the Health and Human Services Analytic Workspace (HHSAW). If you have never set your keyring before and or do not know what this is refering to, just type `keyring::key_set('hhsaw', username = 'ALastname@kingcounty.gov')` into your R console (making sure to replace the username). The default is 'hhsaw'. 

There is no need to specify any or all of the arguments listed above. As the following example shows, the default arguments for `get_population` provide the overall most recent year's estimated King County population.
```{r}
get_population()[]
```

***

## Example analyses
***Note 1**: The use of `head()` below is not necessary. It is a convenience function that displays the first 6 rows of data and was used to keep the output in this vignette tidy.*

***Note 2**: The use of `[]` after get_population() is used to print the output to the console. Typically, you would not print the results but would save them as an object. E.g., `my.pop.est <- get_population()`.*

### Geographic estimates
**WA**
```{r}
get_population(kingco = F, geo_type = "zip")[]
```

**King County**
```{r}
get_population()[]
```

**King County Regions**
```{r}
get_population(geo_type = c("region"),
                   group_by = c("geo_id"))[]
```

**King County HRAs**
```{r}
    head(get_population(geo_type = c("hra"),
                   group_by = c("geo_id"))[])  
```

**King County Zip codes**
```{r}
    head(get_population(geo_type = c("zip"),
                   group_by = c("geo_id"))[])  
```

**King County Census Tracts**
```{r}
    head(get_population(geo_type = c("tract"),
                   group_by = c("geo_id"))[])  
```

**King County Census Block Groups**
```{r}
    head(get_population(geo_type = c("blkgrp"),
                   group_by = c("geo_id"))[])  
```

**King County Census Blocks**
```{r}
    head(get_population(geo_type = c("blk"),
                   group_by = c("geo_id"))[])  
```

### Other simple arguments
**King County multiple years combined**
```{r}
get_population(years = 2017:2019)[]
```

**King County multiple years stratified**
```{r}
get_population(years = 2017:2019, group_by = "years")[]
```

**King County multiple ages combined**
```{r}
get_population(ages = 65:70)[]
```

**King County multiple ages stratified**
```{r}
get_population(ages = 65:70, group_by = "ages")[]
```

**King County female only**
```{r}
get_population(genders = "F")[]
```

**King County gender stratified**
```{r}
get_population(group_by = "genders")[]
```

**King County AIAN-NH**
```{r}
get_population(races = "aian", race_type = "race_eth")[]
```

**King County AIAN regardless of Hispanic ethnicity**
```{r}
get_population(races = "aian", race_type = "race")[]
```

**King County stratified by Hispanic as race**
```{r}
get_population(race_type = "race_eth", group_by = "race_eth")[]
```

**King County stratified by race-NH**
```{r}
get_population(race_type = "race", group_by = "race")[]
```

### Complex arguments
**King County regions stratified by year and gender**
```{r}
get_population(geo_type = "region", years = 2017:2019, group_by = c("geo_id", "years", "genders"))[]
```

**King County regions stratified by year -- Female Hispanic and Asian-NH residents aged 16-25 only**
```{r}
get_population(ages = 16:25, genders = "F", years = 2017:2019, races = c("hispanic", "asian"), geo_type = "region", race_type = "race_eth", group_by = c("geo_id", "years", "race_eth"))[]
```
