---
title: "Leading Causes of Death"
format:
  html:
    embed-resources: true
  gfm:
    prefer-html: false
vignette: >
  %\VignetteIndexEntry{Leading Causes of Death}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{knitr}
editor: source
---

## Six Steps to Identifying the Leading Causes of Death

### Some important notes before we begin

- The CDC's [National Center for Health Statistics (NCHS)](https://www.cdc.gov/nchs/index.htm) collects death certificate data from state vital statistics offices. NCHS performs extensive quality assurance, codes the causes of death using the 10th revision of the International Classification of Diseases (ICD-10), and sends the standardized death data back to state vital statistics offices. They also compile the data to form a comprehensive national database of mortality statistics. Each year they publish an instruction manual with ICD-10 cause of death lists for tabulating mortality statistics, [like this one from 2020](https://www.cdc.gov/nchs/data/dvs/Part9InstructionManual2020-508.pdf). The two key tables for the purposes of public health practice are:

  - **113 Selected Causes of Death, Enterocolitis due to *Clostridium difficile*, and COVID-19,** which is used for the general analysis of mortality and for ranking leading causes of death. This list is typically used for individuals aged 1 year and older. You can view this list with `rads::death_113()` and can generate counts using `rads::death_113_count()`.

  - **130 Selected Causes of Infant Death**, which is used for the analysis of infant mortality and for ranking leading causes of infant death. This list is used for infants, i.e., those who are under 1 year old. You can view this list with `rads::death_130()` and can generate counts using `rads::death_130_count()`.

- Infant death analyses are special.

  - In addition to using a different set of cause of death functions, *infant death analyses do not use standard death certificate & population data*.

    - The ***numerator*** comes from the infant death files. In Washington State, these are downloaded from [secureaccess.wa.gov/doh/chsdatafiles/](https://secureaccess.wa.gov/doh/chsdatafiles/).

    - The ***denominator*** is the number of *live births* recorded in birth certificate data rather than the estimated population with age less than 1 year (though they are very similar).

  - The ***rate*** is *per 1,000* rather than per 100,000.

  - Infant deaths for a given geography include any infant who dies and **either** was born to a mother who resided in that geography at the time of birth **or** was an infant who resided in that geography at the time of death.

  - If infant death data lack geocoded information, you can extract geographic data from birth certificate data by linking with the `birth_cert_encrypt` or by linking to the `state_file_number` in death certificate data.

  - Since the numerator and the denominator come from two different datasets, it is easiest to treat them as counts (infant deaths) and populations (live births) for use with `rads::age_standardize`. Age adjusting for a single age group doesn't make sense, so we use the crude results and confidence intervals.

- By default, [rads death functions](https://github.com/PHSKC-APDE/rads/wiki/death_functions) use the ICD-10 codes in the `underlying_cod_code` in your death data. This is the column that you will want to use for an *underlying cause of death* analysis, which is almost certainly what you will want to do. If you want to change this default, think deeply about your rationale before proceeding.

### Overview

The code below provides an example of how to conduct a leading causes of death analysis using RADS. For the sake of simplicity, we limit the stratification to sex/gender (female & male) for the years 2019 to 2023.

This protocol has six steps:

1.  Set up R

2.  Get data

3.  Get counts

4.  Merge population data onto counts

5.  Calculate age standardized rates

6.  Identify top 10 leading causes based on counts

## (1) Set up R

```{r}
#| warning: false
library(rads)
library(data.table)
```

```{r}
#| echo: false
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

## (2) Get data

### Non-infant deaths (ages ≥ 1)

```{r}
#| warning: false
death <- rads.data::synthetic_death

death <- death[year %in% 2019:2023]

death <- death[, .(year, sex, age, underlying_cod_code)]

# Non-infant deaths only
death <- death[age != 0] 
```

### Infant deaths (ages \< 1)

```{r}
#| warning: false
infdeath <- rads.data::synthetic_infdeath

infdeath <- infdeath[year %in% 2019:2023]

infdeath <- infdeath[, .(year, sex, underlying_cod_code)]

```

### Reference tables

Get *reference tables* with `cause`, `causeid`, and `cause_category` values for infant and non-infant causes of death. These will be used to aggregate causes of death into more meaningful categories.

```{r}
#| warning: false

# Categories of non-infant deaths
death_groups <- rads.data::icd_nchs113causes_raw[, .(cause = leading.cause.group.alias, 
                                                     causeid, 
                                                     cause_category = cause.category)]

# Categories of infant deaths
infdeath_groups <- rads.data::icd_nchs130causes_raw[, .(cause = leading.cause.group.alias, 
                                                        causeid, 
                                                        cause_category = cause.category)]
```

Take a peek at one of the reference tables you just created ...

```{r}
#| echo: true
#| results: 'hide'
head(death_groups)
```

```{r}
#| echo: false
pretty_kable(head(death_groups))
```

As you can see with Tuberculosis in rows 4 & 5, multiple `causeid`s will sometimes be rolled up into one of the `r uniqueN(death_groups$cause)` `cause` values. Therefore the `cause` values allow us to aggregate the `causeid` values into fewer categories.

## (3) Get counts

### Non-infant deaths (ages ≥ 1)

```{r}
#| warning: false
#| message: false
# Do not specify`cause` or `causeids` arguments in order to get all causes of death
death_count <- death_113_count(
  ph.data = death,
  icdcol = 'underlying_cod_code',
  by = c('sex', 'age')
)
```

Tidy the table of non-infant death counts

```{r}
#| warning: false
# Make sure causeid is an integer 
death_count[, causeid := as.integer(causeid)]

# Merge on the leading cause of death group identifier
death_count <- merge(death_count, 
                     death_groups, 
                     by = 'causeid', 
                     all.x= TRUE, 
                     all.y= FALSE)

# Fill in values for 'All causes' and COVID-19
death_count[cause.of.death %in% c('All causes', 'COVID-19 (U07.1)'), 
            cause := cause.of.death]
death_count[cause.of.death == 'All causes', cause_category := cause.of.death]
death_count[cause.of.death == 'COVID-19 (U07.1)', cause_category := 'Infectious disease']

# Delete if missing cause (these are cause.of.death == 'Missing/Unknown')
death_count <- death_count[!is.na(cause)]

# Delete if 'unspecified', 'not elsewhere classified', etc. 
death_count <- death_count[cause != '999']

# Delete if missing sex because cannot merge with population
death_count <- death_count[!is.na(sex)]

# Sum deaths for each unique strata and keep select columns 
# This is necessary because a single `cause` can map to more than one `cause.of.death`
death_count <- death_count[ ,
                            .(deaths = sum(deaths)),
                            .(age,
                              gender = sex, 
                              cause, 
                              cause_category)]
```

Take a peek at the tidied table of non-infant death counts

```{r}
#| echo: true
#| results: 'hide'
head(death_count[deaths >= 10]) 
```

```{r}
#| echo: false
pretty_kable(head(death_count[deaths >= 10]))
```

### Infant deaths (ages\< 1)

```{r}
#| warning: false
# Do not specify`cause` or `causeids` arguments in order to get all causes of death
infdeath_count <- death_130_count(
  ph.data = infdeath,
  icdcol = 'underlying_cod_code',
  by = 'sex'
)
```

Tidy the table of infant death counts

```{r}
#| warning: false
# Make sure causeid is an integer 
infdeath_count[, causeid := as.integer(causeid)]

# Add on the leading cause of death group identifier ----
infdeath_count <- merge(infdeath_count, 
                     infdeath_groups, 
                     by = 'causeid', 
                     all.x= TRUE, all.y= FALSE)

# Fill in values for 'All causes' and COVID-19
infdeath_count[cause.of.death %in% c('All causes', 'COVID-19 (U07.1)'), 
            cause := cause.of.death]
infdeath_count[cause.of.death == 'All causes', cause_category := cause.of.death]
infdeath_count[cause.of.death == 'COVID-19 (U07.1)', cause_category := 'Infectious disease']

# Delete if missing cause (these are cause.of.death == 'Missing/Unknown')
infdeath_count <- infdeath_count[!is.na(cause)]

# Delete if 'unspecified', 'not elsewhere classified', etc. 
infdeath_count <- infdeath_count[cause != '999']

# Delete if missing sex because cannot merge with population
infdeath_count <- infdeath_count[!is.na(sex)]

# Sum deaths for each unique strata and keep select columns 
infdeath_count <- infdeath_count[ , 
                                  .(deaths = sum(deaths), age = 0L), 
                                  .(gender = sex, 
                                    cause, 
                                    cause_category)]

```

Take a peek at the tidied table of infant death counts

```{r}
#| echo: true
#| results: 'hide'
head(infdeath_count[deaths >= 10]) 
```

```{r}
#| echo: false
pretty_kable(head(infdeath_count[deaths >= 10]))
```

## (4) Merge corresponding population onto the count data

### Non-infants (ages ≥ 1)

Be sure that your population data provides the same years and demographic strata that you have in your death counts!

```{r}
#| warning: false
# Get population for the years and demographic strata corresponding to your death counts
pop <- rads.data::synthetic_population

pop <- pop[year %in% 2019:2023 & age %in% 1:100]

pop <- pop[, .(pop = sum(pop)), .(age, gender)]
  
# Merge the population data onto the non-infant death count data
combo <- merge(death_count, 
               pop, 
               by = c('age', 'gender'), 
               all.x= TRUE, 
               all.y= TRUE)

```

### Infants (ages\< 1)

Remember that the infant population should be the number of *live births*, not the estimated population.

```{r}
#| warning: false
# Get infant population from birth data - by sex
infpop <- rads.data::synthetic_birth

infpop <- infpop[year %in% 2019:2023]

infpop <- infpop[, .(pop = .N), .(gender = sex)]

infpop <- infpop[!is.na(gender)]

# Merge the population (live birth) data onto the infant death count data
infcombo <- merge(infdeath_count, 
               infpop, 
               by = 'gender', 
               all.x= TRUE, 
               all.y= TRUE)

```

## (5) Calculate age standardized rates

### Non-infant deaths (ages ≥ 1)

```{r}
#| warning: false
# Non-infant deaths are age standardized
rates <- rads::age_standardize(ph.data = combo, 
                               my.count = 'deaths', 
                               my.pop = 'pop', 
                               by = c('gender', 'cause', 'cause_category')
)

rates <- rates[, .(gender, 
                   ranking = NA_integer_, 
                   cause_category, 
                   cause, 
                   count, 
                   rate = adj.rate,
                   rate_lower = adj.lci,
                   rate_upper = adj.uci)]
```

Take a peek at the non-infant rates ...

```{r}
#| warning: false
#| echo: true
#| results: 'hide'
head(rates[count >= 10]) 
```

```{r}
#| echo: false
pretty_kable(head(rates[count >= 10]))
```

### Infant deaths (ages\< 1)

```{r}
#| warning: false
# Infant deaths are NOT age standardized because there is only one age
infrates <- rads::age_standardize(ph.data = infcombo, 
                                  my.count = 'deaths', 
                                  my.pop = 'pop', 
                                  by = c('gender', 'cause', 'cause_category'),
                                  per = 1000 # 100,000 except for infant mortality
)

infrates <- infrates[, .(gender, 
                         ranking = NA_integer_, 
                         cause_category, 
                         cause, 
                         count, 
                         rate = crude.rate,
                         rate_lower = crude.lci,
                         rate_upper = crude.uci)]
```

Take a peek at the infant rates ...

```{r}
#| warning: false
#| echo: true
#| results: 'hide'
head(infrates[count >= 10]) 
```

```{r}
#| echo: false
pretty_kable(head(infrates[count >= 10]))
```

## (6) Identify leading causes based on COUNTS (using random ties method)

### Non-infant deaths (ages ≥ 1)

```{r}
#| warning: false
# Create ranking (minus one, because want 'All causes' to have a rank of zero)
ranking <- rates[, ranking := frank(-count, ties.method = 'dense')-1, .(gender)]

# Keep top 10 rankings
ranking <- ranking[ranking %in% 0:10]

# Sort the data by gender and ranking
setorder(ranking, gender, ranking)

```

View the top three leading causes of death by gender ...

```{r}
#| echo: true
#| eval: false
# Infant causes of death
print(ranking[count >= 10, .SD[1:4], gender])
```

```{r}
#| warning: false
#| echo: false
#| eval: true
pretty_kable(ranking[count >= 10, .SD[1:4], gender])
```

### Infant deaths (ages\< 1)

```{r}
#| warning: false
# Create ranking
infranking <- infrates[, ranking := frank(-rate, ties.method = 'dense')-1, .(gender)]

# Keep top 10 rankings
infranking <- infranking[ranking %in% 0:10]

# Sort the data by gender and ranking
setorder(infranking, gender, ranking)

```

View the top three leading causes of death by gender ...

```{r}
#| echo: true
#| eval: false
# Infant causes of death
print(infranking[count >= 10, .SD[1:4], gender])
```

```{r}
#| warning: false
#| echo: false
#| eval: true
pretty_kable(infranking[count >= 10, .SD[1:4], gender])
```

## :star2: Finished!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
