---
title: "Utility Functions"
format: 
  gfm:
    prefer-html: false
    self-contained: true
  pdf: default
  html: default
editor: visual
---

# Introduction

The `rads` package includes a suite of small utility functions designed to simplify common data manipulation tasks and accelerate analysis workflows.

The core utility functions we'll cover in this vignette are:

-   [`calc_age()`](#calc_age): Calculate age based on two dates
-   [`convert_to_date()`](#convert_to_date): Convert various formats to Date objects
-   [`create_dictionary()`](#create_dictionary): Generate a comprehensive data dictionary for a data.table or data.frame
-   [`data_modeler()`](#data_modeler): Create uncorrelated data and code that recreates data for sharing publicly
-   [`format_time()`](#format_time-and-format_time_simple): Format numeric values into human-readable time chunks
-   [`format_time_simple()`](#format_time-and-format_time_simple): Format numeric values into a single time chunk
-   [`generate_yaml()`](#generate_yaml): Generate a YAML file for loading R data to SQL
-   [`multi_t_test()`](#multi_t_test): Perform t-tests for Multiple Comparisons with Aggregate Estimates
-   [`round2()`](#round2): Improved rounding function
-   [`string_clean()`](#string_clean): Clean string and factor columns
-   [`std_error()`](#std_error): Calculate standard error of the mean
-   [`substrRight()`](#substrright): Extract substring from the right
-   [`tsql_validate_field_types()`](#tsql_validate_field_types): Validate R data against field types for SQL Server
-   [`tsql_chunk_loader()`](#tsql_chunk_loader): Load large datasets to SQL Server in chunks
-   [`validate_hhsaw_key()`](#validate_hhsaw_key): Validates HHSAW keys and, if possible, opens a connection to HHSAW

Let's explore each of these functions with examples to demonstrate their usage and benefits.


# Setting up the environment

First, let's load the necessary packages:

```{r}
#| warning: false
#| message: false
#| echo: false
library(kableExtra)

# function to beautify tables
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

```{r}
#| warning: false
#| message: false
library(rads)
library(data.table)
```

# Functions

## calc_age() {#calc_age}

The `calc_age()` function calculates age in years based on two dates. This function has identified numerous mistakes in ages provided in DOH vital statistics.

### `calc_age` is better than subtracting the years

```{r}
#| echo: true
#| results: 'hide'
birth_date <- as.Date("1990-08-02")
current_date <- as.Date("2024-08-01")
simple_age <- year(current_date) - year(birth_date)
proper_age <- calc_age(from = birth_date, to = current_date)

head(data.table(birth_date, current_date, simple_age, proper_age))
```

```{r}
#| echo: false
pretty_kable(head(data.table(birth_date, current_date, simple_age, proper_age)))
```

### `calc_age` is better than days / 365.25

```{r}
#| echo: true
#| results: 'hide'
dates <- data.table(
  birth = seq.Date(as.Date('2000-02-25'), as.Date('2000-03-02'), 1), 
  death = seq.Date(as.Date('2024-03-02'), as.Date('2024-02-25'), -1))
dates[, simple_age := round(as.integer(death - birth)/365.25)]
dates[, proper_age := calc_age(birth, death)]

head(dates)
```

```{r}
#| echo: false
pretty_kable(head(dates))
```

## convert_to_date() {#convert_to_date}

`convert_to_date()` attempts to convert various US style date formats to R Dates.

```{r}
#| echo: true
#| results: 'hide'
dates <- data.table(orig = c("2024-07-25", "07/26/2024", 
                             "07-27-2024", "20240728", 
                             "29 July, 2024", "July 30, 2024"))
dates[, rads_dates := convert_to_date(orig)]
dates[, simple_dates := as.Date(orig)]

head(dates)
```

```{r}
#| echo: false
pretty_kable(head(dates))
```

This function is particularly useful when dealing with inconsistent date formats within a single vector or column of your data.

## create_dictionary() {#create_dictionary}

The `create_dictionary()` function generates a comprehensive data dictionary for a given data.table or data.frame. It provides details about variable types, sample values, and can incorporate descriptions and notes from a reference table. The function can be used to both understand a new dataset and document the products of your ETL processes.

### Basic usage

Let's create a sample dataset and generate a dictionary:

```{r}
#| echo: true
#| results: 'hide'
dt <- data.table(
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  gender = factor(c("F", "M", "M", "M", "F")),
  age = c(25, 30, 35, 40, 45),
  score = c(85.5, 92.0, 78.5, 88.0, 95.5)
)

dict <- create_dictionary(ph.data = dt, source = "sample_data")

head(dict, 10)
```

```{r}
#| echo: false
pretty_kable(head(dict, 10))
```

### Suppressing sensitive information

You can suppress sample values for sensitive columns:

```{r}
#| echo: true
#| results: 'hide'
dict_suppressed <- create_dictionary(ph.data = dt, 
                                     source = "sample_data", 
                                     suppress = c("name"))

head(dict_suppressed, 10)
```

```{r}
#| echo: false
pretty_kable(head(dict_suppressed, 10))
```

### Incorporating reference information

You can include additional descriptions and notes using a reference table:

```{r}
#| echo: true
#| results: 'hide'
ref_table <- data.table(
  source = rep("sample_data", 4),
  varname = c("name", "age", "gender", "score"),
  desc = c("Full name", "Age in years", "Gender", "Test score"),
  notes = c("May contain special characters", 
            "Rounded to nearest year", "Binary classification", 
            "Range: 0-100")
)

dict_with_ref <- create_dictionary(ph.data = dt, 
                                   source = "sample_data", 
                                   ph.ref = ref_table)

head(dict_with_ref, 8)
```

```{r}
#| echo: false
pretty_kable(head(dict_with_ref, 8))
```

### Handling numerous unique values

For datasets with many unique values, create_dictionary() provides summaries:

```{r}
#| echo: true
#| results: 'hide'
set.seed(98104)
large_dt <- data.table(
  id = 1:1000,
  category = sample(letters[1:10], 1000, replace = TRUE),
  value = round2(rnorm(1000), 2)
)

large_dict <- create_dictionary(ph.data = large_dt, 
                                source = "large_sample")

head(large_dict, 8)
```

```{r}
#| echo: false
pretty_kable(head(large_dict, 8))
```

## data_modeler() {#data_modeler}

`data_modeler()` receives a data.table object and returns a data.table of similar structure and content, but with uncorrelated data across columns. For example, a column of dates will become a column of dates of a similar range, and distribution, but randomly distributed across observations. Common use cases include:

-   demonstrating a statistic or visualization without exposing the audience to "real" data or results
-   sharing code that generates a data set for minimally viable examples used in requesting help from, or teaching, others
-   creating data sets programmatically for (unit) testing or training

There are two kinds of output that `data_modeler()` returns, either the data table itself, or a list object of strings that can be evaluated to create the data table. In addition, `data_modeler()` can output-to-terminal a human readable version of the generation code that the user can copy and paste into their own scripts for sharing publicly.

`data_modeler()` makes some guesses as to if a column's content should be randomized (such as person names or blood pressures) or kept but randomly distributed (such as levels of response to a question or city names). This is intended to generate random data or to retain original categories where appropriate. It bases these guesses on the data type (integer or string), and complexity (e.g. 75% being unique data), but not the meaning of the data (i.e. it doesn't know that a number is a "zip code" and doesn't know that a string is a "human name"). Variables that are identified as non-categorical will, in most cases, be replaced with a uniform distribution of noisy, but similarly shaped, data (e.g. floating point numbers of the same range and precision, but in a random uniform distribution, or a "lore ipsum" string of similar lengths). Variables identified as categorical will, in most cases, be replaced with identical categories in a similar distribution to the original, but randomly assigned across observations (a data set of 95% males and 5% females would still generate a data set of roughly 95% males and 5% females).

Two parameters are available to adjust the assumptions data_modeler uses to identify categorical values. These are "categorical_threshold" and "proportional_threshold". 

- "categorical_threshold": defines how many unique categories can exist in the variable before data_modeler assumes the variable isn't categorical. Above this number, and the value will not be considered categorical. This currently defaults to 61.
- "proportional_threshold": defines the proportion of values that can be unique before data_modeler() assumes the variable isn't categorical. This currently defaults to 0.75.

### data privacy caveats

#### uncorrelated may not always equal non-informative

`data_modeler()` creates an uncorrelated column of data to substitute the original by randomly reassigning values across that column. However, this does not ensure that meaningful information is not passed along. Note that, in most cases, categorical observations are distributed in proportion to the source data. So, if the provided data set is of categorical outcomes in a single population, the uncorrelated result would still show a similar rate of that outcome in that population because all observations would be for the same population group. The analyst should consider if this is acceptable or not and take measures to correct if necessary by either changing the input data or modifying the synthetic data. A simple approach in the above example would be to add one or more additional population groups so that the outcomes would be distributed removing the ability to estimate the rate for the smaller population (in an uncorrelated result, the rate of the categorical outcome would tend to be similar in all populations.)

### walkthrough

For the examples below, we create a "real" data set and show how to use `data_modeler()` and how the synthetic and "real" data differ. 

Our demo "real" data has one indicator variable that represents an outcome that is never in the White population, sometimes in the Asian population, and always in other Races. In an uncorrelated (and so non-informative) synthetic data set, we would expect this indicator to no longer be biased by race.

```{r create demonstration "real" data}
DT.real.data <- data.table(
   id = 1:1000,
   chi_geo_kc = sample(c('King County',NA), 1000, replace = T),
   chi_race_7 = factor(sample(c("Asian", "AIAN", "Black", "Hispanic", "NHPI", "White", "Other", "Multiple", NA), 1000, replace = T, prob = c(.19,.01,.07,.11,.01,.35,.07,.14,.02)), levels = c("Asian", "AIAN", "Black", "Hispanic", "NHPI", "White", "Other", "Multiple", NA)),
   chi_sex = as.factor(sample(c("Male","Female"), 1000, replace = T)),
   chi_geo_region = factor(sample(c("South", "North", "Seattle", "East"), 1000, replace = T), levels = c("South","North","Seattle","East")))

# Attach a biased indicator.
#indicator1, all White population are "never", all Asians population, "sometimes", and everyone else, "always"
   DT.real.data[, indicator1 := ifelse(chi_race_7 == "White", "never",
                         ifelse(chi_race_7 == "Asian", "sometimes",
                                "always"))]
```

```{r create a data.table of synthetic data}
#| echo: true


##Observe that modeled data have the same structure as the original
DT.synthetic.data <- data_modeler(ph.data = DT.real.data, #our "real" data
                                number_of_observations =  100, #number of observations to create in the synthetic data
                                comments = TRUE, #whether or not to include comments in code (only matters for return_code == TRUE or print_code == TRUE)
                                return_code = FALSE, #return a list object of strings for creating the synthetic data instead of the data itself. leave FALSE to receive synthetic data
                              print_code = FALSE) #print the generation code to terminal. useful for both review and if you wish to copy into other contexts, such as a shareable script or in a chat with a LLM/AI
```

You can observe that the data structures are the same for both data frames:

```{r}
#| echo: fenced
str(DT.real.data)
str(DT.synthetic.data)
```

But we can see that the original bias of White's in the never group, Asians in the sometimes group, and all others in the always group now no longer holds.

```{r}
#| echo: true
prop.table(table(DT.real.data[,.(indicator1,chi_race_7)]))
prop.table(table(DT.synthetic.data[,.(indicator1,chi_race_7)]))
```

Notice that you may see extreme, apparently biased, results in the outcomes. This is still randomly distributed and you can observe these change if you recreate the synthetic.

You can review, share, and execute the code used to create the synthetic data in two ways. Manually by printing it to terminal and copying it into your R script, or by receiving and processing the text object. The first is more useful for review and sharing with others, the later is more useful for programmatic handling of the instruction set.

```{r}
#| echo: true

DT.synthetic.data <- data_modeler(ph.data = DT.real.data,
                                number_of_observations =  100, 
                                comments = TRUE, 
                                return_code = FALSE,
                              print_code = TRUE) # change this to true to copy code from your terminal output
```

The code below is what is used to generate the synthetic data. Using this approach you can see the data, sample approach and rates for each variable. Because we ran `data_modeler()` with `comments = TRUE`, we also see a human-readable string at the end of each line describing how each variable was classified by data_modeler. This allows you to review the logic, make changes, and share the code with others.

```{r code from terminal}
DT <- data.table( 
`id` = as.integer(1:100), # as an identifying (all unique) integer 
`chi_geo_kc` = sample(c(NA, "King County"), 100, replace = TRUE, prob = c(0.494, 0.506)), # as categorical character (non factor) 
`chi_race_7` = factor(sample(c("Black", "White", "AIAN", NA, "Asian", "Multiple", "Other", "NHPI", "Hispanic"), 100, replace = TRUE, prob = c(0.193, 0.014, 0.067, 0.11, 0.012, 0.36, 0.064, 0.157, 0.023)), levels = c("Asian", "AIAN", "Black", "Hispanic", "NHPI", "White", "Other", "Multiple"), ordered = FALSE), # as factor 
`chi_sex` = factor(sample(c("Male", "Female"), 100, replace = TRUE, prob = c(0.518, 0.482)), levels = c("Female", "Male"), ordered = FALSE), # as factor 
`chi_geo_region` = factor(sample(c("Seattle", "South", "North", "East"), 100, replace = TRUE, prob = c(0.241, 0.229, 0.258, 0.272)), levels = c("South", "North", "Seattle", "East"), ordered = FALSE), # as factor 
`indicator1` = sample(c("always", "never", NA, "sometimes"), 100, replace = TRUE, prob = c(0.424, 0.36, 0.193, 0.023))) # as categorical character (non factor)
```

Note, the above approach requires the user to copy and paste output from their terminal. If you wish to check, modify, and extract the code without human intervention, you need to receive the text as a list object using `return_code = TRUE`:

```{r}

DT.synthetic.data.instructions <- data_modeler(ph.data = DT.real.data,
                                number_of_observations =  100, 
                                comments = FALSE, 
                                return_code = TRUE, #setting this to "TRUE" will cause data_modeler() to return a list object of the text instead of a DT of synthetic data
                                print_code = FALSE)

head(DT.synthetic.data.instructions)
```

After processing the text as desired, you can execute it, creating the desired synthetic data object, as follows:

```{r}
## Manually convert received code into data
codeText <- paste(unlist(DT.synthetic.data.instructions), collapse =" \n" )
DT.synthetic.data.manual <- eval( parse(text = paste0(codeText)))
```

Now you can visually confirm that the class structure of each of the tables is the same:

```{r}
str(DT.real.data)
str(DT.synthetic.data)
str(DT.synthetic.data.manual)

sapply(DT.real.data, class)
sapply(DT.synthetic.data, class)
sapply(DT.synthetic.data.manual, class)

```

Finally, each time data_modeler is executed, there is a randomness to how the data are created. If you wish to create identical data sets, be sure to set seed before executing:

```{r}
DT.synthetic.data.random <- data_modeler(ph.data = DT.real.data,
                                number_of_observations =  100, 
                                comments = TRUE, 
                                return_code = FALSE,
                                print_code = FALSE) 

set.seed(1000)
DT.synthetic.data.s1000a <- data_modeler(ph.data = DT.real.data,
                                number_of_observations =  100, 
                                comments = TRUE, 
                                return_code = FALSE,
                                print_code = FALSE) 

set.seed(1000)
DT.synthetic.data.s1000b <- data_modeler(ph.data = DT.real.data,
                                number_of_observations =  100, 
                                comments = TRUE, 
                                return_code = FALSE,
                                print_code = FALSE) 

```

Comparing a table with a random seed to one with a fixed seed, we expect the resulting tables to differ:

```{r}
identical(DT.synthetic.data.random, DT.synthetic.data.s1000a)
```

When using the same seed for distinct rounds, we would expect the resulting tables to be identical:

```{r}
identical(DT.synthetic.data.s1000a, DT.synthetic.data.s1000b)
```

## format_time() and format_time_simple() {#format_time-and-format_time_simple}

These functions format numeric values (including dates) into human-readable time chunks. They are often used for summarizing years and can be used for identifying unexpected gaps.

```{r}
#| echo: true
#| results: 'hide'
years <- c(2001:2003, 
           2005, 
           2008, 
           2011:2012)

dates <- as.Date(c("2024-01-01", "2024-01-02", "2024-01-03",
                   "2024-01-06", 
                   "2024-01-09", "2024-01-10", "2024-01-11", "2024-01-12"))

dt <- data.table(desc = c('format_time', 'format_time_simple', '',
                          'format_time', 'format_time_simple'), 
                 data = c(format_time(years), format_time_simple(years), '', 
                          format_time(dates), format_time_simple(dates)))
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

## generate_yaml() {#generate_yaml}

`generate_yaml()` generates a YAML file for SQL loading based on a data.frame or data.table.

```{r, echo=TRUE, eval=FALSE, results='hide'}
#| echo: true
#| results: 'hide'
data <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25.1, 30.2, 35.3, 40.4, 45.5)
)

yaml_output <- generate_yaml(data, schema = "dbo", table = "example_table", datasource = 'WADOH_CHS')
print(yaml_output)
```

```{r}
#| echo: false
data <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25.1, 30.2, 35.3, 40.4, 45.5)
)

yaml_output <- quiet(generate_yaml(data, schema = "dbo", table = "example_table", datasource = 'WADOH_CHS'))
print(yaml_output)
```

The output can be be used to specify the field_types for the `tsql_validate_field_types()` and `tsql_chunk_loader()` functions described below.

```{r}
#| echo: true
field_types <- unlist(yaml_output$vars)
print(field_types)
```

## multi_t_test() {#multi_t_test}

The `multi_t_test()` function performs t-tests comparing multiple groups against a reference group using means, standard errors, and the sample size (when available).

### Basic usage

Let's create a sample dataset and use `multi_t_test()` to compare birthweights across different maternal age groups:

```{r}
#| echo: true
#| results: 'hide'
# Sample data
age_groups <- c("18-24", "25-29", "30-34", "35-39", "40+")
birthweight_means <- c(3150, 3450, 3400, 3250, 3100)  # in grams
birthweight_ses <- c(50, 45, 40, 55, 60)
sample_sizes <- c(500, 800, 750, 400, 200)
reference_group <- 3  # comparing all groups to the 30-34 age group

# Perform multi_t_test
birthweight_comparison <- multi_t_test(means = birthweight_means,
                                       ses = birthweight_ses,
                                       reference_index = reference_group,
                                       n = sample_sizes,
                                       alternative = 'two.sided',
                                       df_method = "estimated", 
                                       alpha = 0.05)

# Add age group labels to the results
birthweight_comparison[, Age_Group := age_groups]

head(birthweight_comparison)
```

```{r}
#| echo: false
pretty_kable(head(birthweight_comparison))
```

### Key features

1.  **Flexible degrees of freedom calculation**: The function offers four methods for calculating degrees of freedom:

    -   "**estimated**" (default): Uses the Welch–Satterthwaite equation
    -   "**conservative**": Uses the minimum possible degrees of freedom (df = 2); lower Type I Error (false +) and higher Type II Error (false -)
    -   "**moderate**": Uses the number of groups minus 1 (df = k - 1)
    -   "**liberal**": Assumes infinite degrees of freedom (df = Inf); higher Type 1 Error (false +) and lower Type II Error (false -)

2.  **Sample size estimation**: If sample sizes are not provided, the function can estimate them based on the distribution of mean values.

3.  **Adjustment for multiple comparisons (optional)**: If desired, it will adjust p-values for multiple comparisons using one of two methods (Holm-Bonferroni or Benjamini-Hochberg).

4.  **Comprehensive output**: The function returns a data.table containing comparison results, including difference in means, confidence intervals, p-values, and significance indicators.

## round2() {#round2}

`round2()` rounds numbers the way we learned in elementary school: if the digit to the right of the rounding position is 5 or greater, round up; otherwise, round down. ***Note!*** *Base R's `round` function follows a different logic (IEC 60559 / IEEE 754 standards), which specifies rounding to the nearest even number.*

```{r}
#| echo: true
#| results: 'hide'
set.seed(98104)
dt <- data.table(orig = seq(0.5, 5.5, 1))
dt[, round2 := round2(orig)]
dt[, baseR := round(orig)]
dt[round2 != baseR, different := '*']
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

Like the base R `round` function, `round2` can also be used to round large numbers to convey uncertainty in estimates

```{r}
# round to hundreds
round2(123456, -2)
# round to thousands
round2(123456, -3)
```

## string_clean() {#string_clean}

`string_clean()` cleans string and factor columns in a data.frame or data.table. It eliminates random white spaces and pseudo-white spaces and sets blanks to true NA.

```{r}
#| echo: true
# create table 
dt <- data.table(
  name = c(" John   ", "Jane  Doe", "Bob Smith", "  ", "  Edward  Jenner"),
  category = c("A ", "   B", "", "D", "E      ")
)
print(dt$category)
```

```{r}
#| echo: true
#| results: 'hide'
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

```{r}
#| echo: true
# clean table
string_clean(dt)
print(dt$category)
```

```{r}
#| echo: true
#| results: 'hide'
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

## std_error() {#std_error}

`std_error()` calculates the standard error of the mean for a vector, columns in a table, or a list. R does not have a built in standard error of the mean function, so we created one for `rads`.

### Use with a vector

```{r}
#| echo: true

set.seed(98104)
x <- rnorm(1000)
se <- std_error(x)
print(paste("Standard Error:", round2(se, 5)))
```

### Use with a data.table / data.frame

```{r}
#| echo: true
set.seed(98104)
dt <- data.table(x = rnorm(1000), y = rnorm(1000))
se <- std_error(dt)
se <- round2(se, 5)
print(se)
```

### Use with a list

```{r}
#| echo: true
set.seed(98104)
mylist <- list(x = rnorm(1000), y = rnorm(1000))
se <- std_error(mylist) # returns a list with the standard errors
se <- lapply(se, round2, n = 5)
print(se)
```

## substrRight() {#substrright}

`substrRight()` extracts a substring from the right side of a string. It complements the base R function `substr`.

```{r}
#| echo: true

text <- "Hello, World!"
substr_right <- substrRight(text, 2, 6)
print(substr_right)
```

This function is useful when you need to extract the right hand portions of strings, e.g., the last 4 digits of a phone number.

## tsql_validate_field_types() {#tsql_validate_field_types}

`tsql_validate_field_types()` validates whether a named vector of TSQL data types is compatible with a data.table. This function is useful when you have a yaml file specifying the TSQL field types to be used when uploading your data to a server.

### A successful validation

```{r}
#| echo: true
#| results: 'hide'
dt <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25, 30, 35, 40, 45), 
  time = Sys.time()
)

field_types <- c(id = "int", 
                 name = "varchar(50)", 
                 age = "int", 
                 time = 'datetime')
validation_result <- tsql_validate_field_types(dt, field_types)
head(validation_result)
```

```{r}
#| echo: false
pretty_kable(head(validation_result))
```

### A failed validation

```{r}
#| echo: true
#| error: true
dt <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = as.character(c(25, 30, 35, 40, 45)), 
  time = Sys.Date()
)

field_types <- c(id = "int", 
                 name = "varchar(50)", 
                 age = "int", # should generate an error
                 time = 'datetime') # should generate an error
validation_result <- tsql_validate_field_types(dt, field_types)
```

## tsql_chunk_loader() {#tsql_chunk_loader}

`tsql_chunk_loader()` loads large datasets to Microsoft SQL Server in 'chunks'. It is suitable for datasets too large for single upload. For extreme sizes try using [bcp](https://learn.microsoft.com/en-us/sql/tools/bcp-utility?view=sql-server-ver16&tabs=windows).

Benefits include:

-   Loads large datasets in chunks, preventing memory and batch size issues
-   Reduces risk of network timeouts during data transfer
-   Attempts multiple uploads per chunk for error resilience
-   Provides progress tracking
-   Allows SQL data type specification and validation
-   Verifies complete data upload

```{r}
#| echo: true

 mydt = data.table(col1 = 1:10000L,  # create integer
                   col2 = 1:10000/3) # create float
 mydt[, col3 := as.Date(Sys.Date()) - col1] # create date
 mydt[, col4 := as.character(col3)] # create string
 
 myfieldtypes <- c(col1 = 'int', col2 = 'float', 
                   col3 = 'date', col4 = 'nvarchar(20)')

 tsql_chunk_loader(
   ph.data = mydt,
   db_conn = rads::validate_hhsaw_key(), # connect to Azure 16 hhs_analytics_workspace
   chunk_size = 3333,
   schema_name = Sys.getenv("USERNAME"),
   table_name = 'JustTesting',
   overwrite = TRUE,
   append = FALSE,
   field_types = myfieldtypes,
   validate_field_types = TRUE, # uses tsql_validate_field_types()
   validate_upload = TRUE
 )
```

## validate_hhsaw_key() {#validate_hhsaw_key}

`validate_hhsaw_key()` validates HHSAW keys **and** opens a connection to HHSAW when possible.

```{r}
#| echo: true
 myConnection <- validate_hhsaw_key(hhsaw_key = 'hhsaw')
 print(myConnection)
```

# Conclusion

These utility functions provide tools for common data manipulation and analysis tasks. By leveraging these functions, you can streamline your workflow, improve code readability, and handle various data-related challenges more efficiently.

Remember to consult the individual function documentation for more detailed information on usage and parameters. Happy coding!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'), ' (rads v', packageVersion('rads'), ')')`*
