---
title: "Utility Functions"
format: 
  gfm:
    prefer-html: false
    self-contained: true
  pdf: default
  html: default
editor: visual
---

# Introduction

The `rads` package includes a suite of small utility functions designed to simplify common data manipulation tasks and accelerate analysis workflows.

The core utility functions we'll cover in this vignette are:

-   [`calc_age()`](#calc_age): Calculate age based on two dates
-   [`convert_to_date()`](#convert_to_date): Convert various formats to Date objects
-   [`create_dictionary()`](#create_dictionary): Generate a comprehensive data dictionary for a data.table or data.frame
-   [`format_time()`](#format_time-and-format_time_simple): Format numeric values into human-readable time chunks
-   [`format_time_simple()`](#format_time-and-format_time_simple): Format numeric values into a single time chunk
-   [`generate_yaml()`](#generate_yaml): Generate a YAML file for loading R data to SQL
-   [`multi_t_test()`](#multi_t_test): Perform t-tests for Multiple Comparisons with Aggregate Estimates
-   [`round2()`](#round2): Improved rounding function
-   [`string_clean()`](#string_clean): Clean string and factor columns
-   [`std_error()`](#std_error): Calculate standard error of the mean
-   [`substrRight()`](#substrright): Extract substring from the right
-   [`tsql_validate_field_types()`](#tsql_validate_field_types): Validate R data against field types for SQL Server
-   [`tsql_chunk_loader()`](#tsql_chunk_loader): Load large datasets to SQL Server in chunks
-   [`validate_hhsaw_key()`](#validate_hhsaw_key): Validates HHSAW keys and, if possible, opens a connection to HHSAW

Let's explore each of these functions with examples to demonstrate their usage and benefits.

# Setting up the environment

First, let's load the necessary packages:

```{r}
#| warning: false
#| message: false
#| echo: false
library(kableExtra)

# function to beautify tables
pretty_kable <- function(dt) { 
  knitr::kable(dt, format = 'markdown')
}
```

```{r}
#| warning: false
#| message: false
library(rads)
library(data.table)
```

## calc_age() {#calc_age}

The `calc_age()` function calculates age in years based on two dates. This function has identified numerous mistakes in ages provided in DOH vital statistics.

### `calc_age` is better than subtracting the years

```{r}
#| echo: true
#| results: 'hide'
birth_date <- as.Date("1990-08-02")
current_date <- as.Date("2024-08-01")
simple_age <- year(current_date) - year(birth_date)
proper_age <- calc_age(from = birth_date, to = current_date)

head(data.table(birth_date, current_date, simple_age, proper_age))
```

```{r}
#| echo: false
pretty_kable(head(data.table(birth_date, current_date, simple_age, proper_age)))
```

### `calc_age` is better than days / 365.25

```{r}
#| echo: true
#| results: 'hide'
dates <- data.table(
  birth = seq.Date(as.Date('2000-02-25'), as.Date('2000-03-02'), 1), 
  death = seq.Date(as.Date('2024-03-02'), as.Date('2024-02-25'), -1))
dates[, simple_age := round(as.integer(death - birth)/365.25)]
dates[, proper_age := calc_age(birth, death)]

head(dates)
```

```{r}
#| echo: false
pretty_kable(head(dates))
```

## convert_to_date() {#convert_to_date}

`convert_to_date()` attempts to convert various US style date formats to R Dates.

```{r}
#| echo: true
#| results: 'hide'
dates <- data.table(orig = c("2024-07-25", "07/26/2024", 
                             "07-27-2024", "20240728", 
                             "29 July, 2024", "July 30, 2024"))
dates[, rads_dates := convert_to_date(orig)]
dates[, simple_dates := as.Date(orig)]

head(dates)
```

```{r}
#| echo: false
pretty_kable(head(dates))
```

This function is particularly useful when dealing with inconsistent date formats within a single vector or column of your data.

## create_dictionary() {#create_dictionary}

The `create_dictionary()` function generates a comprehensive data dictionary for a given data.table or data.frame. It provides details about variable types, sample values, and can incorporate descriptions and notes from a reference table. The function can be used to both understand a new dataset and document the products of your ETL processes.

### Basic usage

Let's create a sample dataset and generate a dictionary:

```{r}
#| echo: true
#| results: 'hide'
dt <- data.table(
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  gender = factor(c("F", "M", "M", "M", "F")),
  age = c(25, 30, 35, 40, 45),
  score = c(85.5, 92.0, 78.5, 88.0, 95.5)
)

dict <- create_dictionary(ph.data = dt, source = "sample_data")

head(dict, 10)
```

```{r}
#| echo: false
pretty_kable(head(dict, 10))
```

### Suppressing sensitive information

You can suppress sample values for sensitive columns:

```{r}
#| echo: true
#| results: 'hide'
dict_suppressed <- create_dictionary(ph.data = dt, 
                                     source = "sample_data", 
                                     suppress = c("name"))

head(dict_suppressed, 10)
```

```{r}
#| echo: false
pretty_kable(head(dict_suppressed, 10))
```

### Incorporating reference information

You can include additional descriptions and notes using a reference table:

```{r}
#| echo: true
#| results: 'hide'
ref_table <- data.table(
  source = rep("sample_data", 4),
  varname = c("name", "age", "gender", "score"),
  desc = c("Full name", "Age in years", "Gender", "Test score"),
  notes = c("May contain special characters", 
            "Rounded to nearest year", "Binary classification", 
            "Range: 0-100")
)

dict_with_ref <- create_dictionary(ph.data = dt, 
                                   source = "sample_data", 
                                   ph.ref = ref_table)

head(dict_with_ref, 8)
```

```{r}
#| echo: false
pretty_kable(head(dict_with_ref, 8))
```

### Handling numerous unique values

For datasets with many unique values, create_dictionary() provides summaries:

```{r}
#| echo: true
#| results: 'hide'
set.seed(98104)
large_dt <- data.table(
  id = 1:1000,
  category = sample(letters[1:10], 1000, replace = TRUE),
  value = round2(rnorm(1000), 2)
)

large_dict <- create_dictionary(ph.data = large_dt, 
                                source = "large_sample")

head(large_dict, 8)
```

```{r}
#| echo: false
pretty_kable(head(large_dict, 8))
```

## format_time() and format_time_simple() {#format_time-and-format_time_simple}

These functions format numeric values (including dates) into human-readable time chunks. They are often used for summarizing years and can be used for identifying unexpected gaps.

```{r}
#| echo: true
#| results: 'hide'
years <- c(2001:2003, 
           2005, 
           2008, 
           2011:2012)

dates <- as.Date(c("2024-01-01", "2024-01-02", "2024-01-03",
                   "2024-01-06", 
                   "2024-01-09", "2024-01-10", "2024-01-11", "2024-01-12"))

dt <- data.table(desc = c('format_time', 'format_time_simple', '',
                          'format_time', 'format_time_simple'), 
                 data = c(format_time(years), format_time_simple(years), '', 
                          format_time(dates), format_time_simple(dates)))
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

## generate_yaml() {#generate_yaml}

`generate_yaml()` generates a YAML file for SQL loading based on a data.frame or data.table.

```{r, echo=TRUE, eval=FALSE, results='hide'}
#| echo: true
#| results: 'hide'
data <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25.1, 30.2, 35.3, 40.4, 45.5)
)

yaml_output <- generate_yaml(data, schema = "dbo", table = "example_table", datasource = 'WADOH_CHS')
print(yaml_output)
```

```{r}
#| echo: false
data <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25.1, 30.2, 35.3, 40.4, 45.5)
)

yaml_output <- quiet(generate_yaml(data, schema = "dbo", table = "example_table", datasource = 'WADOH_CHS'))
print(yaml_output)
```

The output can be be used to specify the field_types for the `tsql_validate_field_types()` and `tsql_chunk_loader()` functions described below.

```{r}
#| echo: true
field_types <- unlist(yaml_output$vars)
print(field_types)
```

## multi_t_test() {#multi_t_test}

The `multi_t_test()` function performs t-tests comparing multiple groups against a reference group using means, standard errors, and the sample size (when available).

### Basic usage

Let's create a sample dataset and use `multi_t_test()` to compare birthweights across different maternal age groups:

```{r}
#| echo: true
#| results: 'hide'
# Sample data
age_groups <- c("18-24", "25-29", "30-34", "35-39", "40+")
birthweight_means <- c(3150, 3450, 3400, 3250, 3100)  # in grams
birthweight_ses <- c(50, 45, 40, 55, 60)
sample_sizes <- c(500, 800, 750, 400, 200)
reference_group <- 3  # comparing all groups to the 30-34 age group

# Perform multi_t_test
birthweight_comparison <- multi_t_test(means = birthweight_means,
                                       ses = birthweight_ses,
                                       reference_index = reference_group,
                                       n = sample_sizes,
                                       alternative = 'two.sided',
                                       df_method = "estimated", 
                                       alpha = 0.05)

# Add age group labels to the results
birthweight_comparison[, Age_Group := age_groups]

head(birthweight_comparison)
```

```{r}
#| echo: false
pretty_kable(head(birthweight_comparison))
```

### Key features

1.  **Flexible degrees of freedom calculation**: The function offers four methods for calculating degrees of freedom:

    -   "**estimated**" (default): Uses the Welch–Satterthwaite equation
    -   "**conservative**": Uses the minimum possible degrees of freedom (df = 2); lower Type I Error (false +) and higher Type II Error (false -)
    -   "**moderate**": Uses the number of groups minus 1 (df = k - 1)
    -   "**liberal**": Assumes infinite degrees of freedom (df = Inf); higher Type 1 Error (false +) and lower Type II Error (false -)

2.  **Sample size estimation**: If sample sizes are not provided, the function can estimate them based on the distribution of mean values.

3.  **Adjustment for multiple comparisons (optional)**: If desired, it will adjust p-values for multiple comparisons using one of two methods (Holm-Bonferroni or Benjamini-Hochberg).

4.  **Comprehensive output**: The function returns a data.table containing comparison results, including difference in means, confidence intervals, p-values, and significance indicators.

## round2() {#round2}

`round2()` rounds numbers the way we learned in elementary school: if the digit to the right of the rounding position is 5 or greater, round up; otherwise, round down. ***Note!*** *Base R's `round` function follows a different logic (IEC 60559 / IEEE 754 standards), which specifies rounding to the nearest even number.*

```{r}
#| echo: true
#| results: 'hide'
set.seed(98104)
dt <- data.table(orig = seq(0.5, 5.5, 1))
dt[, round2 := round2(orig)]
dt[, baseR := round(orig)]
dt[round2 != baseR, different := '*']
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

Like the base R `round` function, `round2` can also be used to round large numbers to convey uncertainty in estimates

```{r}
# round to hundreds
round2(123456, -2)
# round to thousands
round2(123456, -3)
```

## string_clean() {#string_clean}

`string_clean()` cleans string and factor columns in a data.frame or data.table. It eliminates random white spaces and pseudo-white spaces and sets blanks to true NA.

```{r}
#| echo: true
# create table 
dt <- data.table(
  name = c(" John   ", "Jane  Doe", "Bob Smith", "  ", "  Edward  Jenner"),
  category = c("A ", "   B", "", "D", "E      ")
)
print(dt$category)
```

```{r}
#| echo: true
#| results: 'hide'
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

```{r}
#| echo: true
# clean table
string_clean(dt)
print(dt$category)
```

```{r}
#| echo: true
#| results: 'hide'
head(dt)
```

```{r}
#| echo: false
pretty_kable(head(dt))
```

## std_error() {#std_error}

`std_error()` calculates the standard error of the mean for a vector, columns in a table, or a list. R does not have a built in standard error of the mean function, so we created one for `rads`.

### Use with a vector

```{r}
#| echo: true

set.seed(98104)
x <- rnorm(1000)
se <- std_error(x)
print(paste("Standard Error:", round2(se, 5)))
```

### Use with a data.table / data.frame

```{r}
#| echo: true
set.seed(98104)
dt <- data.table(x = rnorm(1000), y = rnorm(1000))
se <- std_error(dt)
se <- round2(se, 5)
print(se)
```

### Use with a list

```{r}
#| echo: true
set.seed(98104)
mylist <- list(x = rnorm(1000), y = rnorm(1000))
se <- std_error(mylist) # returns a list with the standard errors
se <- lapply(se, round2, n = 5)
print(se)
```

## substrRight() {#substrright}

`substrRight()` extracts a substring from the right side of a string. It complements the base R function `substr`.

```{r}
#| echo: true

text <- "Hello, World!"
substr_right <- substrRight(text, 2, 6)
print(substr_right)
```

This function is useful when you need to extract the right hand portions of strings, e.g., the last 4 digits of a phone number.

## tsql_validate_field_types() {#tsql_validate_field_types}

`tsql_validate_field_types()` validates whether a named vector of TSQL data types is compatible with a data.table. This function is useful when you have a yaml file specifying the TSQL field types to be used when uploading your data to a server.

### A successful validation

```{r}
#| echo: true
#| results: 'hide'
dt <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = c(25, 30, 35, 40, 45), 
  time = Sys.time()
)

field_types <- c(id = "int", 
                 name = "varchar(50)", 
                 age = "int", 
                 time = 'datetime')
validation_result <- tsql_validate_field_types(dt, field_types)
head(validation_result)
```

```{r}
#| echo: false
pretty_kable(head(validation_result))
```

### A failed validation

```{r}
#| echo: true
#| error: true
dt <- data.table(
  id = 1:5,
  name = c("Alicia", "Bianca", "Clarence", "Dwane", "Eve"),
  age = as.character(c(25, 30, 35, 40, 45)), 
  time = Sys.Date()
)

field_types <- c(id = "int", 
                 name = "varchar(50)", 
                 age = "int", # should generate an error
                 time = 'datetime') # should generate an error
validation_result <- tsql_validate_field_types(dt, field_types)
```

## tsql_chunk_loader() {#tsql_chunk_loader}

`tsql_chunk_loader()` loads large datasets to Microsoft SQL Server in 'chunks'. It is suitable for datasets too large for single upload. For extreme sizes try using [bcp](https://learn.microsoft.com/en-us/sql/tools/bcp-utility?view=sql-server-ver16&tabs=windows).

Benefits include:

-   Loads large datasets in chunks, preventing memory and batch size issues
-   Reduces risk of network timeouts during data transfer
-   Attempts multiple uploads per chunk for error resilience
-   Provides progress tracking
-   Allows SQL data type specification and validation
-   Verifies complete data upload

```{r}
#| echo: true

 mydt = data.table(col1 = 1:10000L,  # create integer
                   col2 = 1:10000/3) # create float
 mydt[, col3 := as.Date(Sys.Date()) - col1] # create date
 mydt[, col4 := as.character(col3)] # create string
 
 myfieldtypes <- c(col1 = 'int', col2 = 'float', 
                   col3 = 'date', col4 = 'nvarchar(20)')

 tsql_chunk_loader(
   ph.data = mydt,
   db_conn = rads::validate_hhsaw_key(), # connect to Azure 16 hhs_analytics_workspace
   chunk_size = 3333,
   schema_name = Sys.getenv("USERNAME"),
   table_name = 'JustTesting',
   overwrite = TRUE,
   append = FALSE,
   field_types = myfieldtypes,
   validate_field_types = TRUE, # uses tsql_validate_field_types()
   validate_upload = TRUE
 )
```

## validate_hhsaw_key() {#validate_hhsaw_key}

`validate_hhsaw_key()` validates HHSAW keys **and** opens a connection to HHSAW when possible.

```{r}
#| echo: true
 myConnection <- validate_hhsaw_key(hhsaw_key = 'hhsaw')
 print(myConnection)
```

# Conclusion

These utility functions provide tools for common data manipulation and analysis tasks. By leveraging these functions, you can streamline your workflow, improve code readability, and handle various data-related challenges more efficiently.

Remember to consult the individual function documentation for more detailed information on usage and parameters. Happy coding!

-- *`r paste0('Updated ', format(Sys.time(), '%B %d, %Y'))`*
